---
title: "HL_DE_analysis-iTALK_visualization"
author: "Stacy Hung"
date: "18/08/2020"
output: html_document
---

## Visualize interactions 

```{r}
library(iTALK)
library(dplyr)

comm_list <- c('growth factor','other','cytokine','checkpoint')
cell_col <- structure(c('#4a84ad','#e874bf','#b79eed', '#ff636b', '#9ef49a', '#4a1dc6'), 
                      names = c("CTL", "Helper T cells", "Naïve T cells", "T regs", "Macrophages", "Proliferation (?)"))

# read in our expression data, filtered for macrophages and T cell clusters
data <- read.table("~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/hl_combined.norm_expr.macrophage_t_cell_clusters.annotated.txt", sep="\t", header = TRUE, stringsAsFactors = FALSE)
colnames(data)[colnames(data) == 'hl.both.combined.celltype'] <- 'cell_type'
colnames(data)[colnames(data) == 'hl.both.combined.case.type'] <- 'diag_rel'
colnames(data)[colnames(data) == 'hl.both.combined.earlyRelapse'] <- 'diag_early_rel'
data$diag_early_rel <- NULL
data$diag_rel <- NULL

# filter for genes of interest:
genes.keep <- c("CXCL10", "CXCR3", "MIF", "CXCR4", "cell_type") # missing CXCL12
data.GOI <- data[genes.keep]

# remove LSARP-15 barcodes
data.minus_outlier <- data.GOI[ !grepl("LSARP-15", rownames(data.GOI)), ]
```

## Get top expressed genes

```{r}
# find top 50 percent highly expressed genes
highly_exprs_genes <- rawParse(data.minus_outlier, top_genes = 100, stats='mean')
expr.CXC_genes <- highly_exprs_genes
```

## Visualization of interactions based on normalized expression (not DE)

```{r}
par(mfrow = c(1,2))
res <- NULL

# NB: CXC receptor-ligand pairs only produce results for the cytokine communication type, so we don't
#     to iterate through all the communication types (e.g. growth facter, etc.)
comm_type <- 'cytokine'
res <- FindLR(expr.CXC_genes, datatype = 'mean count', comm_type = comm_type)
res <- res[ order(res$cell_from_mean_exprs*res$cell_to_mean_exprs, decreasing = T), ]

NetView(res_cat, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5)
LRPlot(res_cat, datatype = 'mean count', cell_col = cell_col, 
       link.arr.lwd = res_cat$cell_from_mean_exprs,
       link.arr.width = res_cat$cell_to_mean_exprs)
```

## Visualize differential expression for just CXC* genes within macrophages and T cell subtypes

```{r}
library(tidyr)

# function that filters for cell type and gene and returns ligand-receptor interactions in given DE dataset
# default is to filter for macrophage, proliferation and T cell subtypes and filter for CXC genes
getLR <- function(de.df) {
  #split "comparison" column by "_" and take the first split element to be cell_type
  de.df <- separate(data = de.df, col = comparison, into = c("cell_type", "group1"), sep = "\\_")

  keeps <- c("gene", "cell_type", "avg_logFC", "p_val", "p_val_adj")
  de.df <- de.df[ keeps ]
  
  keeps.rename <- c("gene", "cell_type", "logFC", "p.value", "q.value")
  colnames(de.df) <- keeps.rename
  
  # filter for cell types of interest
  cell_types.keep <- c("CTL", "Helper T cells", "Macrophages", "Naïve T cells", "T regs", "Proliferation (?)")
  de.cell_types <- subset(de.df, de.df$cell_type %in% cell_types.keep)
  
  # filter for genes of interest
  de.GOI <- de.cell_types[grep("CXC", de.cell_types$gene), ]
  
  # only cytokine communications exist between CXC components
  comm_type <- "cytokine"
  res <- FindLR(de.GOI, datatype='DEG', comm_type = comm_type)
  res <- res[order(abs(res$cell_from_logFC*res$cell_to_logFC), decreasing = TRUE), ]

  return(res)
}

# diagonstic vs. relapse
de_results.diag_v_rel <- read.table("~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/de_analysis/by_cluster/minus_LSARP-15/de.diag_vs_rel.clusters.txt", sep = "\t", header = TRUE)
res <- getLR(de_results.diag_v_rel)
LRPlot(res, datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res$cell_from_logFC, link.arr.width = res$cell_to_logFC)
NetView(res, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5)

# diagnostic vs. early relapse
de_results.diag_v_early_rel <- read.table("~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/de_analysis/by_cluster/minus_LSARP-15/de.diag_vs_early_relapse.clusters.txt", sep = "\t", header = TRUE)
res <- getLR(de_results.diag_v_early_rel)
LRPlot(res, datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res$cell_from_logFC, link.arr.width = res$cell_to_logFC)
NetView(res, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5)

# diagnostic vs. late relapse
de_results.diag_v_late_rel <- read.table("~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/de_analysis/by_cluster/minus_LSARP-15/de.diag_vs_late_relapse.clusters.txt", sep = "\t", header = TRUE)
res <- getLR(de_results.diag_v_late_rel)
LRPlot(res, datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res$cell_from_logFC, link.arr.width = res$cell_to_logFC)
NetView(res, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5)

# diagnostic vs. refractory
de_results.diag_v_ref <- read.table("~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/de_analysis/by_cluster/minus_LSARP-15/de.diag_vs_refractory.clusters.txt", sep = "\t", header = TRUE)
res <- getLR(de_results.diag_v_ref)
LRPlot(res, datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res$cell_from_logFC, link.arr.width = res$cell_to_logFC)
NetView(res, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5, vertex.size = 30, margin = 0, 
edge.curved = 0)


```

## Differential expression - approach 1: with iTALK - currently buggy

```{r}
# Find DEGenes of regulatory T cells* and macrophages between these 2 groups.
# Here T cells refers to one of: Naive T cells, CTL, Treg, and helper T cells.
# Important! The contrasts refers to the two different group (values) -- in our case 2 = diagnostic and 1 = relapse

# make sure there is only 1 extra columns at the end: cell_type and compare_group
data$compare_group <- data$diag_rel
# keep for future reference:
diag_rel.df <- data$diag_rel
diag_early_rel.df <- data$diag_early_rel
# remove extra columns
data$diag_rel <- NULL
data$diag_early_rel <- NULL
data$barcode <- NULL

# pre-filter data to avoid having to doing it on the spot for every DE analysis:
data.CTL <- filter(data, data$cell_type == 'CTL')
data.helperTcells <- filter(data, data$cell_type == 'Helper T cells')
data.naiveTcells <- filter(data, data$cell_type == 'Naïve T cells')
data.Tregs <- filter(data, data$cell_type == 'T regs')
data.macrophage <- filter(data, data$cell_type == 'Macrophage')

deg.ctl <- DEG(data.CTL, method='MAST', c('diagnostic', 'relapse'))
deg.macrophage <- DEG(data.macrophage, method='MAST', contrast = c('diagnostic', 'relapse'))
```

## Differential expression - approach 2: from Seurat

Make sure results are in data frame with columns: gene, cell_type, logFC, p.value, and q.value

```{r}
de_results.diag_v_rel <- read.table("~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/de_analysis/by_cluster/de.diag_vs_rel.clusters.txt", sep = "\t", header = TRUE)

keeps <- c("gene", "cell_type", "avg_logFC", "p_val", "p_val_adj")
keeps.rename <- c("gene", "cell_type", "logFC", "p.value", "q.value")

# extract results within macrophage cluster
de_results.diag_v_rel.macrophage <- de_results.diag_v_rel[grep("Macrophages", de_results.diag_v_rel$comparison), ]
de_results.diag_v_rel.macrophage$cell_type <- "macrophage"
de_results.diag_v_rel.macrophage$comparison <- NULL
de_results.diag_v_rel.macrophage$pct.1 <- NULL
de_results.diag_v_rel.macrophage$pct.2 <- NULL
de_results.diag_v_rel.macrophage <- de_results.diag_v_rel.macrophage[ keeps ]
colnames(de_results.diag_v_rel.macrophage) <- keeps.rename

# extract results within T-cell cluster (start with CTL)
de_results.diag_v_rel.CTL <- de_results.diag_v_rel[grep("CTL", de_results.diag_v_rel$comparison), ]
de_results.diag_v_rel.CTL$cell_type <- "CTL"
de_results.diag_v_rel.CTL$comparison <- NULL
de_results.diag_v_rel.CTL$pct.1 <- NULL
de_results.diag_v_rel.CTL$pct.2 <- NULL
de_results.diag_v_rel.CTL <- de_results.diag_v_rel.CTL[ keeps ]
colnames(de_results.diag_v_rel.CTL) <- keeps.rename

# extract results within T-cell cluster (Helper T cells)
de_results.diag_v_rel.helperTCells <- de_results.diag_v_rel[grep("Helper T cells", de_results.diag_v_rel$comparison), ]
de_results.diag_v_rel.helperTCells$cell_type <- "Helper T cells"
de_results.diag_v_rel.helperTCells$comparison <- NULL
de_results.diag_v_rel.helperTCells$pct.1 <- NULL
de_results.diag_v_rel.helperTCells$pct.2 <- NULL
de_results.diag_v_rel.helperTCells <- de_results.diag_v_rel.helperTCells[ keeps ]
colnames(de_results.diag_v_rel.helperTCells) <- keeps.rename

# extract results within T-cell cluster (Naive T cells)
de_results.diag_v_rel.naiveTCells <- de_results.diag_v_rel[grep("Naïve T cells", de_results.diag_v_rel$comparison), ]
de_results.diag_v_rel.naiveTCells$cell_type <- "Naïve T cells"
de_results.diag_v_rel.naiveTCells$comparison <- NULL
de_results.diag_v_rel.naiveTCells$pct.1 <- NULL
de_results.diag_v_rel.naiveTCells$pct.2 <- NULL
de_results.diag_v_rel.naiveTCells <- de_results.diag_v_rel.naiveTCells[ keeps ]
colnames(de_results.diag_v_rel.naiveTCells) <- keeps.rename

# extract results within T-cell cluster (T regs)
de_results.diag_v_rel.TRegs <- de_results.diag_v_rel[grep("T regs", de_results.diag_v_rel$comparison), ]
de_results.diag_v_rel.TRegs$cell_type <- "T regs"
de_results.diag_v_rel.TRegs$comparison <- NULL
de_results.diag_v_rel.TRegs$pct.1 <- NULL
de_results.diag_v_rel.TRegs$pct.2 <- NULL
de_results.diag_v_rel.TRegs <- de_results.diag_v_rel.TRegs[ keeps ]
colnames(de_results.diag_v_rel.TRegs) <- keeps.rename


de_results.diag_v_early_rel <- read.table("~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/de_analysis/by_cluster/de.diag_vs_early_relapse.clusters.txt", sep = "\t", header = TRUE)

keeps <- c("gene", "cell_type", "avg_logFC", "p_val", "p_val_adj")
keeps.rename <- c("gene", "cell_type", "logFC", "p.value", "q.value")

# extract results within macrophage cluster
de_results.diag_v_early_rel.macrophage <- de_results.diag_v_early_rel[grep("Macrophages", de_results.diag_v_early_rel$comparison), ]
de_results.diag_v_early_rel.macrophage$cell_type <- "macrophage"
de_results.diag_v_early_rel.macrophage$comparison <- NULL
de_results.diag_v_early_rel.macrophage$pct.1 <- NULL
de_results.diag_v_early_rel.macrophage$pct.2 <- NULL
de_results.diag_v_early_rel.macrophage <- de_results.diag_v_early_rel.macrophage[ keeps ]
colnames(de_results.diag_v_early_rel.macrophage) <- keeps.rename

# extract results within T-cell cluster (start with CTL)
de_results.diag_v_early_rel.CTL <- de_results.diag_v_early_rel[grep("CTL", de_results.diag_v_early_rel$comparison), ]
de_results.diag_v_early_rel.CTL$cell_type <- "CTL"
de_results.diag_v_early_rel.CTL$comparison <- NULL
de_results.diag_v_early_rel.CTL$pct.1 <- NULL
de_results.diag_v_early_rel.CTL$pct.2 <- NULL
de_results.diag_v_early_rel.CTL <- de_results.diag_v_early_rel.CTL[ keeps ]
colnames(de_results.diag_v_early_rel.CTL) <- keeps.rename

# extract results within T-cell cluster (Helper T cells)
de_results.diag_v_early_rel.helperTCells <- de_results.diag_v_early_rel[grep("Helper T cells", de_results.diag_v_early_rel$comparison), ]
de_results.diag_v_early_rel.helperTCells$cell_type <- "Helper T cells"
de_results.diag_v_early_rel.helperTCells$comparison <- NULL
de_results.diag_v_early_rel.helperTCells$pct.1 <- NULL
de_results.diag_v_early_rel.helperTCells$pct.2 <- NULL
de_results.diag_v_early_rel.helperTCells <- de_results.diag_v_early_rel.helperTCells[ keeps ]
colnames(de_results.diag_v_early_rel.helperTCells) <- keeps.rename

# extract results within T-cell cluster (Naive T cells)
de_results.diag_v_early_rel.naiveTCells <- de_results.diag_v_early_rel[grep("Naïve T cells", de_results.diag_v_early_rel$comparison), ]
de_results.diag_v_early_rel.naiveTCells$cell_type <- "Naïve T cells"
de_results.diag_v_early_rel.naiveTCells$comparison <- NULL
de_results.diag_v_early_rel.naiveTCells$pct.1 <- NULL
de_results.diag_v_early_rel.naiveTCells$pct.2 <- NULL
de_results.diag_v_early_rel.naiveTCells <- de_results.diag_v_early_rel.naiveTCells[ keeps ]
colnames(de_results.diag_v_early_rel.naiveTCells) <- keeps.rename

# extract results within T-cell cluster (T regs)
de_results.diag_v_early_rel.TRegs <- de_results.diag_v_early_rel[grep("T regs", de_results.diag_v_early_rel$comparison), ]
de_results.diag_v_early_rel.TRegs$cell_type <- "T regs"
de_results.diag_v_early_rel.TRegs$comparison <- NULL
de_results.diag_v_early_rel.TRegs$pct.1 <- NULL
de_results.diag_v_early_rel.TRegs$pct.2 <- NULL
de_results.diag_v_early_rel.TRegs <- de_results.diag_v_early_rel.TRegs[ keeps ]
colnames(de_results.diag_v_early_rel.TRegs) <- keeps.rename

```

## iTALK: find significant ligand-receptor pairs and do the plotting

```{r}
par(mfrow = c(1,2)) # create two panels for plotting

de_results.Tcell <- de_results.diag_v_rel.CTL
# de_results.Tcell <- de_results.diag_v_rel.helperTCells
# de_results.Tcell <- de_results.diag_v_rel.naiveTCells
# de_results.Tcell <- de_results.diag_v_rel.TRegs

# de_results.Tcell <- de_results.diag_v_early_rel.CTL
# de_results.Tcell <- de_results.diag_v_early_rel.helperTCells
# de_results.Tcell <- de_results.diag_v_early_rel.naiveTCells
de_results.Tcell <- de_results.diag_v_early_rel.TRegs


res <- NULL # create an empty results dataframe

for(comm_type in comm_list) {
  res_cat<-FindLR(de_results.Tcell, de_results.diag_v_early_rel.macrophage, datatype='DEG',comm_type=comm_type)
  res<-rbind(res,res_cat)
}
res <- na.omit(res)
res<-res[order(abs(res$cell_from_logFC*res$cell_to_logFC),decreasing=T),]

# plots
LRPlot(res, datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res$cell_from_logFC,
           link.arr.width = res$cell_to_logFC)
NetView(res, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5)
  # title(comm_type)
```

```{r}
# !!! buggy !!!
  
for(comm_type in comm_list) {
  res_cat <- FindLR(de_results.diag_v_rel.macrophage, de_results.diag_v_rel.CTL, 
                    datatype = 'DEG', comm_type = comm_type)
  res_cat <- res_cat[ order(res_cat$cell_from_logFC*res_cat$cell_to_logFC, decreasing = T), ]
  
  # plot by ligand category
  if(nrow(res_cat) == 0) {
    next
  } else if(nrow(res_cat >= 20)) {
    LRPlot(res_cat[1:20, ], datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res_cat$cell_from_logFC[1:20],
           link.arr.width = res_cat$cell_to_logFC[1:20])
  } else {
    LRPlot(res_cat, datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res_cat$cell_from_logFC, 
           link.arr.width = res_cat$cell_to_logFC)
  }
  NetView(res_cat, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5)
  title(comm_type)
  res <- rbind(res, res_cat)
}

if(is.null(res)) {
  print('No significant pairs found')
} else if(nrow(res) >= 20) {
  res <- res[order(res$cell_from_logFC*res$cell_to_logFC, decreasing = T), ][1:20, ]
  NetView(res, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5)
  LRPlot(res[1:20, ], datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res$cell_from_logFC[1:20],
         link.arr.width = res$cell_to_logFC[1:20])
} else {
  NetView(res, col = cell_col, vertex.label.cex = 1, arrow.width = 1, edge.max.width = 5)
  LRPlot(res, datatype = 'DEG', cell_col = cell_col, link.arr.lwd = res$cell_from_logFC, 
         link.arr.width = res$cell_to_logFC)
}
```
