---
title: "LSARP_analysis-CD4_CTL_clustering"
author: "Stacy Hung"
date: "20/01/2021"
output: html_document
---

## Load libraries

```{r}
library(Seurat)
library(dplyr)
library(pheatmap)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(ggrepel)
library(ggpubr)
```

## Step 1: Filter "CD4" based on expression / non-expression of marker combinations

```{r}
hl_all.combined <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.annotated.rds")
  
# Step 1: Filter cells for "CD4+" based on:
#   Expression of CD3 (CD3D), CD4
#   No expression of CD56 (NCAM1), CD8 (CD8A), CD19, and CD68
hl.ctl_CD3_CD4_pos.cells <- which(hl_all.combined@assays$RNA@data["CD3D",] > 0 & 
                               hl_all.combined@assays$RNA@data["CD4",] > 0 &
                               hl_all.combined@assays$RNA@data["NCAM1",] == 0 &
                               hl_all.combined@assays$RNA@data["CD8A",] == 0 &
                               hl_all.combined@assays$RNA@data["CD19",] == 0 &
                               hl_all.combined@assays$RNA@data["CD68",] == 0
                            ) # 24596 (15% of all cells)
hl.filtered.CD4_markers <- subset(hl_all.combined, cells = hl.ctl_CD3_CD4_pos.cells)
# what proportion of cells belong to EBV+ patients? (12%)
# NEG     POS     NA
# 19313   2641   102
```

## Step 2: Subcluster and plot heatmaps of clusters

```{r}
DefaultAssay(hl.filtered.CD4_markers) <- "integrated"
hl.filtered.CD4_markers <- FindNeighbors(hl.filtered.CD4_markers, reduction = "pca", dims = 1:30)
hl.filtered.CD4_markers.subclustered <- FindClusters(hl.filtered.CD4_markers, resolution = 0.4)
Idents(hl.filtered.CD4_markers.subclustered) <- hl.filtered.CD4_markers.subclustered$seurat_clusters

# plot umap
pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/CD4/umap-CD4_pos-subclusters.pdf", width = 6, height = 5)
DimPlot(hl.filtered.CD4_markers.subclustered, reduction = "umap", label = TRUE) 
dev.off()

# plot average expression heatmap for component markers
avg_exp.heatmaps <- avgComponentExpressionHeatmap(data = hl.filtered.CD4_markers.subclustered)

# plot individual expression heatmaps for each cluster
printSCHeatmapsPerCluster(data = hl.filtered.CD4_markers.subclustered, 
                          plot_folder = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/CD4/")
```

## Step 3: Annotate clusters (Tomo)

```{r}
# attach manual annotations
hl.filtered.CD4_markers.subclustered <- RenameIdents(hl.filtered.CD4_markers.subclustered, 
                                                     `9` = "CD4 CTL")
hl.filtered.CD4_markers.subclustered$celltype2 <- Idents(hl.filtered.CD4_markers.subclustered)

pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/CD4/umap-CD4-subclusters-annotated.pdf", width = 6, height = 5)
DimPlot(hl.filtered.CD4_markers.subclustered, reduction = "umap", label = TRUE) 
dev.off()

saveRDS(hl.filtered.CD4_markers.subclustered, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/CD4_CTL/hl.filtered.CD4_markers.subclustered.annotated.rds")
```

## Compare proportion of CD4 CTL cluster:

To calculate proportion for each (patient): Total CD4+ CTL cells / [ Total CD4+ cells | Total cells in patient ]

- between patients

```{r}
# get cells belonging to "CD4 CTL" cluster (and have EBV status)
hl.CD4_CTL.cells <- which(hl.filtered.CD4_markers.subclustered$celltype2 == "CD4 CTL" &
                            hl.filtered.CD4_markers.subclustered$ebv %in% c("POS", "NEG"))
hl.CD4_CTL <- subset(hl.filtered.CD4_markers.subclustered, cells = hl.CD4_CTL.cells)

# get df containing cell barcode and ebv status
#hl.CD4_CTL.ebv <- cbind(as.data.frame(hl.CD4_CTL$orig.ident), as.data.frame(hl.CD4_CTL$ebv))

# plot proportion of all cells that are CD4+ CTL in each patient
hl.CD4_CTL.patient.ebv <- as.data.frame(table(hl.CD4_CTL$orig.ident))
colnames(hl.CD4_CTL.patient.ebv) <- c("sample_id", "num_cells")

# map total CD4+ cells (before subclustering)
hl.filtered.CD4.patient <- as.data.frame(table(hl.filtered.CD4_markers$orig.ident))
colnames(hl.filtered.CD4.patient) <- c("sample_id", "total_CD4_cells")

# map total cells per patient
hl_all.patient <- as.data.frame(table(hl_all.combined$orig.ident))
colnames(hl_all.patient) <- c("sample_id", "total_cells")

# merge CD4+ CTL df with total cells dfs
hl.CD4_CTL.df1 <- merge(hl.CD4_CTL.patient.ebv, hl.filtered.CD4.patient, by = "sample_id")
hl.CD4_CTL.df <- merge(hl.CD4_CTL.df1, hl_all.patient, by = "sample_id")

# calculate proportion of CD4+ CTL cells (based on total CD4 cells or total cells)
hl.CD4_CTL.df <- transform(hl.CD4_CTL.df, proportion_CD4 = (num_cells / total_CD4_cells) * 100 )
hl.CD4_CTL.df <- transform(hl.CD4_CTL.df, proportion_total = (num_cells / total_cells) * 100 )

# map on EBV status
final.cohort <- read.table("~/Documents/sc_HL_relapse/data/final_cohort.plus_RLN.assay_version.txt", sep = "\t", header = TRUE, fill = TRUE)
final.cohort <- filter(final.cohort, final.cohort$FinalCohort == "yes")
hl.CD4_CTL.df$ebv <- plyr::mapvalues(
  x = hl.CD4_CTL.df$sample_id,
  from = as.character(final.cohort$Sample),
  to = as.character(final.cohort$EBVStatus)
)

# plot proportions based on total CD4 filteered cells as denominator
p1 <- ggplot(data=hl.CD4_CTL.df, aes(x=reorder(sample_id, -proportion_CD4), y=proportion_CD4, fill=ebv)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  xlab("") + ylab("CD4+ CTL cells (Cluster 9) / CD4 filtered cells (%)") +
  labs(fill = "EBV")

# plot proportions based on total cells as denominator
p2 <- ggplot(data=hl.CD4_CTL.df, aes(x=reorder(sample_id, -proportion_total), y=proportion_total, fill=ebv)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  xlab("") + ylab("CD4+ CTL cells (Cluster 9) / Total cells (%)") +
  labs(fill = "EBV")

pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/CD4/barplot-CD4_CTL_proportions_by_patient.pdf", 
    width = 10, height = 9.5)
grid.arrange(p1, p2)
dev.off()
```

Compare CD4 CTL proportions (relative to CD4 filtered cells and relative to total cells) for:

- diagnostic vs. *all* relapse
- diagnostic vs. refractory / early relapse
- HL (diagnostic + relapse) vs. RLN

For these comparisons, we do *not* care about EBV status

```{r}
hl.CD4_CTL.cells <- which(hl.filtered.CD4_markers.subclustered$celltype2 == "CD4 CTL")
hl.CD4_CTL <- subset(hl.filtered.CD4_markers.subclustered, cells = hl.CD4_CTL.cells)
hl.CD4_CTL.patient.ebv <- as.data.frame(table(hl.CD4_CTL$orig.ident))
colnames(hl.CD4_CTL.patient.ebv) <- c("sample_id", "num_cells")
hl.filtered.CD4.patient <- as.data.frame(table(hl.filtered.CD4_markers$orig.ident))
colnames(hl.filtered.CD4.patient) <- c("sample_id", "total_CD4_cells")
hl_all.patient <- as.data.frame(table(hl_all.combined$orig.ident))
colnames(hl_all.patient) <- c("sample_id", "total_cells")
hl.CD4_CTL.df1 <- merge(hl.CD4_CTL.patient.ebv, hl.filtered.CD4.patient, by = "sample_id")
hl.CD4_CTL.df <- merge(hl.CD4_CTL.df1, hl_all.patient, by = "sample_id")
hl.CD4_CTL.df <- transform(hl.CD4_CTL.df, proportion_CD4 = (num_cells / total_CD4_cells) * 100 )
hl.CD4_CTL.df <- transform(hl.CD4_CTL.df, proportion_total = (num_cells / total_cells) * 100 )

# map on relapse and early relapse status and biology
hl.CD4_CTL.df$case.type <- plyr::mapvalues(
  x = hl.CD4_CTL.df$sample_id,
  from = as.character(final.cohort$Sample),
  to = as.character(final.cohort$case.type)
)
hl.CD4_CTL.df$EarlyRelapse <- plyr::mapvalues(
  x = hl.CD4_CTL.df$sample_id,
  from = as.character(final.cohort$Sample),
  to = as.character(final.cohort$EarlyRelapse)
)
hl.CD4_CTL.df$Biology <- plyr::mapvalues(
  x = hl.CD4_CTL.df$sample_id,
  from = as.character(final.cohort$Sample),
  to = as.character(final.cohort$Biology)
)

# calculate mean CD4 CTL proportion per case type

# diagnostic vs. relapse - relative to CD4 filtered cells
hl.CD4_CTL.diag_vs_rel <- as.data.frame(hl.CD4_CTL.df %>%
  select(case.type, proportion_CD4, proportion_total) %>%
  group_by(case.type) %>%
  mutate(mean_proportion_CD4 = mean(proportion_CD4)))
# diagnostic vs. relapse - relative to all cells
hl.CD4_CTL.diag_vs_rel <- as.data.frame(hl.CD4_CTL.diag_vs_rel %>%
  group_by(case.type) %>%
  mutate(mean_proportion_total = mean(proportion_total)))

# diagnostic vs. earlyRelapse - relative to CD4 filtered cells
hl.CD4_CTL.diag_vs_earlyRel <- as.data.frame(hl.CD4_CTL.df %>%
  select(EarlyRelapse, proportion_CD4, proportion_total) %>%
  group_by(EarlyRelapse) %>%
  mutate(mean_proportion_CD4 = mean(proportion_CD4)))
# diagnostic vs. earlyRelapse - relative to all cells
hl.CD4_CTL.diag_vs_earlyRel <- as.data.frame(hl.CD4_CTL.diag_vs_earlyRel %>%
  group_by(EarlyRelapse) %>%
  mutate(mean_proportion_total = mean(proportion_total)))

# HL vs. RLN - relative to CD4 filtered cells
hl.CD4_CTL.hl_vs_rln <- as.data.frame(hl.CD4_CTL.df %>%
  select(Biology, proportion_CD4, proportion_total) %>%
  group_by(Biology) %>%
  mutate(mean_proportion_CD4 = mean(proportion_CD4)))
# HL vs. RLN - relative to all cells
hl.CD4_CTL.hl_vs_rln <- as.data.frame(hl.CD4_CTL.hl_vs_rln %>%
  group_by(Biology) %>%
  mutate(mean_proportion_total = mean(proportion_total)))

# violin plots

# Diagnostic vs. relapse: filter diagnostic / relapse cases
hl.CD4_CTL.diag_vs_rel <- subset(hl.CD4_CTL.diag_vs_rel, 
                                 hl.CD4_CTL.diag_vs_rel$case.type %in% c("diagnostic", "relapse"))

p.diag_vs_rel1 <- ggplot(hl.CD4_CTL.diag_vs_rel, aes(x = case.type, y = proportion_CD4)) + 
  geom_violin(aes(fill = mean_proportion_CD4), trim = TRUE, scale = "width") +
  stat_compare_means(comparisons = list(c("diagnostic", "relapse")), label = "p.format") +
  geom_boxplot(width=0.2) +
  xlab("") + ylab("") +
  scale_y_continuous(trans='log2')

p.diag_vs_rel2 <- ggplot(hl.CD4_CTL.diag_vs_rel, aes(x = case.type, y = proportion_total)) + 
  geom_violin(aes(fill = mean_proportion_total), trim = TRUE, scale = "width") +
  stat_compare_means(comparisons = list(c("diagnostic", "relapse")), label = "p.format") +
  geom_boxplot(width=0.2) +
  xlab("") + ylab("") +
  scale_y_continuous(trans='log2')
#stat_summary(fun.data = custom_quantile, geom = "boxplot", width=0.2) +

pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/CD4/violin-CD4_CTL-diag_vs_rel.pdf", 
    width = 12, height = 5)
grid.arrange(p.diag_vs_rel1, p.diag_vs_rel2, ncol=2)
dev.off()

# Diagnostic vs. early relapse: filter for diagnostic / early relapse cases
hl.CD4_CTL.diag_vs_earlyRel <- subset(hl.CD4_CTL.diag_vs_earlyRel, 
                                      hl.CD4_CTL.diag_vs_earlyRel$EarlyRelapse == "diagnostic" |
                                        hl.CD4_CTL.diag_vs_earlyRel$EarlyRelapse == "earlyRelapse")

p.diag_vs_earlyRel1 <- ggplot(hl.CD4_CTL.diag_vs_earlyRel, aes(x = EarlyRelapse, y = proportion_CD4)) + 
  geom_violin(aes(fill = mean_proportion_CD4), trim = TRUE, scale = "width") +
  stat_compare_means(comparisons = list(c("diagnostic", "earlyRelapse")), label = "p.format") +
  geom_boxplot(width=0.2) +
  xlab("") + ylab("") +
  scale_y_continuous(trans='log2')

p.diag_vs_earlyRel2 <- ggplot(hl.CD4_CTL.diag_vs_earlyRel, aes(x = EarlyRelapse, y = proportion_total)) + 
  geom_violin(aes(fill = mean_proportion_total), trim = TRUE, scale = "width") +
  stat_compare_means(comparisons = list(c("diagnostic", "earlyRelapse")), label = "p.format") +
  geom_boxplot(width=0.2) +
  xlab("") + ylab("") +
  scale_y_continuous(trans='log2')

pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/CD4/violin-CD4_CTL-diag_vs_earlyRel.pdf", 
    width = 12, height = 5)
grid.arrange(p.diag_vs_earlyRel1, p.diag_vs_earlyRel2, ncol=2)
dev.off()

# HL vs. RLN
p.hl_vs_rln1 <- ggplot(hl.CD4_CTL.hl_vs_rln, aes(x = Biology, y = proportion_CD4)) + 
  geom_violin(aes(fill = mean_proportion_CD4), trim = TRUE, scale = "width") +
  stat_compare_means(comparisons = list(c("HL", "RLN")), label = "p.format") +
  geom_boxplot(width=0.2) +
  xlab("") + ylab("") +
  scale_y_continuous(trans='log2')

p.hl_vs_rln2 <- ggplot(hl.CD4_CTL.hl_vs_rln, aes(x = Biology, y = proportion_total)) + 
  geom_violin(aes(fill = mean_proportion_total), trim = TRUE, scale = "width") +
  stat_compare_means(comparisons = list(c("HL", "RLN")), label = "p.format") +
  geom_boxplot(width=0.2) +
  xlab("") + ylab("") +
  scale_y_continuous(trans='log2')

pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/CD4/violin-CD4_CTL-HL_vs_RLN.pdf", 
    width = 12, height = 5)
grid.arrange(p.hl_vs_rln1, p.hl_vs_rln2, ncol=2)
dev.off()


```


## Function: Create heatmap of average expression of clusters across component markers

Parameters:
  - data: Seurat object with $seurat_clusters

```{r}
avgComponentExpressionHeatmap <- function(data) {
  # get component markers
  component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", 
                                   sep = "\t", header = TRUE, fill = TRUE)
  # first, get list of highly-variable genes
  hvg <- data@assays$integrated@var.features
  # only look at markers that are in the HVG list
  markers.to.plot <- subset(hvg, hvg %in% component.mappings$Gene)
  # get average expression of each gene
  cluster.averages <- AverageExpression(data, features = markers.to.plot, return.seurat = TRUE)
  #data.avg <- GetAssayData(object = cluster.averages, slot = "scale.data")
  data.avg <- GetAssayData(object = cluster.averages, slot = "data", assay = "SCT")
  # now match component gene list to those being plotted
  component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
  rownames(component.mappings) <- component.mappings$Gene
  # order by component then gene name so heatmap legend makes sense
  components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
  component.mappings$Gene <- NULL
  component.mappings$Alt_name <- NULL
  # reorder expression matrix to match component mappings
  data.avg.ordered <- data.avg[match(rownames(components.ordered), rownames(data.avg)), ]

  heatmaps <- vector('list', 2)
  
  # heatmap of average expression per cluster across all component genes
  heatmaps[[1]] <- pheatmap(data.avg.ordered, 
         annotation_row = component.mappings,
         cluster_rows = FALSE, 
         annotation_names_row = FALSE, 
         angle_col = 45, 
         gaps_row = c(14,20,22,24,27,32,34,38,39,40,41,44,51,53))
  # rotated version of heatmap
  heatmaps[[2]] <- pheatmap(t(data.avg.ordered), 
         annotation_col = component.mappings,
         cluster_cols = FALSE, 
         annotation_names_col = FALSE, 
         gaps_col = c(14,20,22,24,27,32,34,38,39,40,41,44,51,53), 
         angle_col = 45)
  
  return(heatmaps)
}

```

## Produce single-cell level heatmaps of expression for EACH cluster

Parameters:
  - data: Seurat object with $seurat_clusters
  - plot_folder: absolute path of folder to print heatmaps to

```{r}
printSCHeatmapsPerCluster <- function(data, plot_folder) {
  # get component markers
  component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", 
                                   sep = "\t", header = TRUE, fill = TRUE)
  # first, get list of highly-variable genes
  hvg <- data@assays$integrated@var.features
  # only look at markers that are in the HVG list
  markers.to.plot <- subset(hvg, hvg %in% component.mappings$Gene)
  # get total number of clusters
  NUM_CLUSTERS = length(unique(data$seurat_clusters))
  # get expression matrix (features by barcodes)
  data.expr <- data@assays$SCT@data
  # retrieve expression for each cell across the component genes of interest
  data.expr.markers <- data.expr[ rownames(data.expr) %in% markers.to.plot , ]
  # match component gene list to those that can be plotted
  component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
  # order by component then gene name so that the heatmap legend makes sense
  rownames(component.mappings) <- component.mappings$Gene
  components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
  components.ordered$Gene <- NULL
  components.ordered$Alt_name <- NULL
  # reorder expression matrix to match component mappings
  data.expr.markers.ordered <- data.expr.markers[ row.names(components.ordered), ]
  # to create separate expr matrices for each cluster, need mappings for cell --> cluster
  data.clusters <- data$seurat_clusters
  # convert to char
  clusters.char <- levels(data.clusters)[data.clusters] # ordered vector of cluster ids (0 through 26)
  barcodes.cluster <- attributes(data.clusters)$names # ordered vector of barcodes (76,607 unique entries)
  clusters.df <- data.frame(barcodes.cluster, clusters.char)
 
  expr.markers <- vector('list', NUM_CLUSTERS)
  heatmaps <- vector('list', NUM_CLUSTERS)
  hl.seurat <- vector('list', NUM_CLUSTERS)
  
  for (c in 1:NUM_CLUSTERS) {
    hl.seurat[[c]] <- subset(data, idents = c-1) # filter by cluster
    hl.seurat[[c]] <- subset(hl.seurat[[c]], features = markers.to.plot) # filter by genes
    
    # retrieve expression for each cell across the component genes of interest
    expr.temp <- hl.seurat[[c]]@assays$SCT@data
    expr.markers[[c]] <- expr.temp[ rownames(expr.temp) %in% markers.to.plot , ]
    
    # reorder expression matrix to match component mappings
    expr.markers[[c]] <- expr.markers[[c]][ row.names(components.ordered), ]
  
    filename <- paste(plot_folder, "heatmap-cluster", 
                      c-1, ".pdf", sep = "")
    pdf(filename, width = 8, height = 8)
    heatmaps[[c]] <- pheatmap(expr.markers[[c]], 
           annotation_row = components.ordered, 
           cluster_rows = FALSE, 
           annotation_names_row = FALSE, 
           angle_col = 45, 
           gaps_row = c(14,20,22,24,27,32,34,38,39,40,41,44,51,53), 
           border_color = NA, show_colnames = FALSE, scale = "none")
    dev.off()
  }
 
}
```
