---
title: "Seurat_analysis"
author: "Stacy Hung"
date: "02/12/2019"
output: html_document
---

```{r}
library(Seurat) # remotes::install_github("satijalab/seurat", ref = "release/4.0.0")
library(SeuratData)
library(SeuratWrappers)
library(sctransform)
library(future) # for parallelization
library(tidyverse)
library(dplyr)
library(tidyr)
library(grid)
library(gridExtra)
library(ggplot2)
library(pheatmap)
library(cowplot)
library(patchwork) # for saving R objects to and loading from *.RDS file
library(scran)
library(batchelor)
library(harmony)
library(SingleR)
library(celldex)
library(MAST) # required for DE analysis using MAST option in Seurat FindMarkers
library(DESeq2)
```

## Examine QC metrics

```{r}
final.cohort <- read.table("~/Documents/sc_HL_relapse/data/final_cohort.plus_RLN.assay_version.txt", sep = "\t", header = TRUE, fill = TRUE)
final.cohort <- filter(final.cohort, final.cohort$FinalCohort == "yes")

rln.metrics <- read.table("~/Documents/sc_HL_relapse/metrics/RLN_metrics.txt", sep = "\t", header = TRUE)
relapse.metrics <- read.table("~/Documents/sc_HL_relapse/metrics/all.relapse.metrics.txt", sep = "\t", header = TRUE)
primary.metrics <- read.table("~/Documents/sc_HL_relapse/metrics/all.primary.metrics.txt", sep = "\t", header = TRUE)

# Liz's favourite metrics:
  # 1. cells recovered --> "Estimated.Number.of.Cells"
  # 2. mean reads per cell --> "Mean.Reads.per.Cell" 
  # 3. median genes per cell --> "Median.Genes.per.Cell"
  # 4. % sequencing saturation --> "Sequencing.Saturation"

# Other relevant metrics:
  # "Fraction.Reads.in.Cells"
  # "Total.Genes.Detected"

keeps <- c("SampleID", "Estimated.Number.of.Cells", "Mean.Reads.per.Cell", "Median.Genes.per.Cell", "Sequencing.Saturation", "Fraction.Reads.in.Cells", "Total.Genes.Detected")
relapse.metrics <- relapse.metrics[c(keeps)]
primary.metrics <- primary.metrics[c(keeps)]
rln.metrics <- rln.metrics[c(keeps)]

# combine datasets
all.metrics <- rbind(primary.metrics, relapse.metrics, rln.metrics)

colnames(all.metrics) <- c("SampleID", "Est. Number of Cells", "Mean Reads / Cell", "Median Genes / Cell", "Sequencing Saturation (%)", "Fraction Reads in Cells (%)", "Total Genes Detected")

# filter out excluded samples
all.metrics <- filter(all.metrics, all.metrics$SampleID %in% final.cohort$Sample)

# map values for chemistry
all.metrics$chemistry <- plyr::mapvalues(
  x = all.metrics$SampleID,
  from = as.character(final.cohort$Sample),
  to = as.character(final.cohort$X3_prime_assay)
)

# convert from wide to long
metrics.long <- gather(all.metrics, metric, value, "Est. Number of Cells":"Total Genes Detected")

# sorted version
data.sorted <- metrics.long %>%
  ungroup() %>%
  arrange(metric, value) %>%
  dplyr::mutate(order = row_number())

# colors: diagnostic - royal blue; relapse - light red; RLN - light purple
p.sorted <- ggplot(data.sorted, aes(x = order, y = value, fill=chemistry)) + 
      geom_bar(stat = "identity", width = 0.8) + 
      facet_wrap( ~ metric, scales = "free") + 
      xlab("") + ylab("") + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 3)) +
      theme(panel.spacing = unit(1, "lines")) +
      scale_fill_manual(values=c("cornflowerblue", "indianred3")) +
      scale_x_continuous(
        breaks = data.sorted$order,
        labels = data.sorted$SampleID,
        expand = c(0,0)
      )

```

## Read in 10X data

Note that 10X has an aggr function that combines multiple samples into a single count matrix, but normalization is performed using downsampling, which is less ideal than using scTransform in Seurat


```{r}
ids.rln <- filter(final.cohort, final.cohort$Cohort == "RLN")$Sample
ids.primary <- filter(final.cohort, final.cohort$Cohort == "Primary")$Sample
ids.relapse <- filter(final.cohort, final.cohort$Cohort == "Relapse")$Sample

# read in RLN data
dataset_loc <- "~/Documents/sc_HL_relapse/cellranger-3.1.0/RLN"
d10x.data <- sapply(ids.rln, function(i) {
  d10x <- Read10X(file.path(dataset_loc, i, ""))
  colnames(d10x) <- paste(sapply(strsplit(colnames(d10x), split="-"), '[[', 1L), i, sep="_")
                          d10x
})
experiment.data <- do.call("cbind", d10x.data)
rln <- CreateSeuratObject(
  experiment.data,
  project = "RLN",
  min.cells = 3, # keep all genes that are expressed in at least 3 cells --> might be too lenient for scTransform
  min.features = 200, # keep all cells that are expressing at least 200 transcripts
  names.field = 2, # the sample id is in the 2nd column after applying the delimiter "_"
  names.delim = "\\_" # the delimiter
) # 20211 features across 15771 samples within 1 assay 

# read in primary data
dataset_loc <- "~/Documents/sc_HL_relapse/cellranger-3.1.0/primary"
d10x.data <- sapply(ids.primary, function(i) {
  d10x <- Read10X(file.path(dataset_loc, i, ""))
  colnames(d10x) <- paste(sapply(strsplit(colnames(d10x), split="-"), '[[', 1L), i, sep="_")
                          d10x
})
experiment.data <- do.call("cbind", d10x.data)
primary <- CreateSeuratObject(
  experiment.data,
  project = "primary",
  min.cells = 3, # keep all genes that are expressed in at least 3 cells --> might be too lenient for scTransform
  min.features = 200, # keep all cells that are expressing at least 200 transcripts
  names.field = 2, # the sample id is in the 2nd column after applying the delimiter "_"
  names.delim = "\\_" # the delimiter
) # 21457 features across 78584 samples within 1 assay 

# read in relapse data
dataset_loc <- "~/Documents/sc_HL_relapse/cellranger-3.1.0/relapse"
d10x.data <- sapply(ids.relapse, function(i) {
  d10x <- Read10X(file.path(dataset_loc, i, ""))
  colnames(d10x) <- paste(sapply(strsplit(colnames(d10x), split="-"), '[[', 1L), i, sep="_")
                          d10x
})
experiment.data <- do.call("cbind", d10x.data)
relapse <- CreateSeuratObject(
  experiment.data,
  project = "relapse",
  min.cells = 3, # keep all genes that are expressed in at least 3 cells --> might be too lenient for scTransform
  min.features = 200, # keep all cells that are expressing at least 200 transcripts
  names.field = 2, # the sample id is in the 2nd column after applying the delimiter "_"
  names.delim = "\\_" # the delimiter
) # 23064 features across 103522 samples within 1 assay


# merge all datasets
hl.merged <- merge(x = rln, y = c(primary, relapse)) # 25299 features across 197,877 samples within 1 assay
```

## Add metadata and basic filters

```{r}
# calculate % mitochondria for each cell
hl.merged[["percent.mt"]] <- PercentageFeatureSet(hl.merged, pattern = "^MT-")

# map metadata
hl.merged.idents <- as.data.frame(hl.merged$orig.ident)
hl.merged.names <- as.character(hl.merged.idents$`hl.merged$orig.ident`)

# case type
hl.merged$case.type <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$case.type))
# cohort
hl.merged$cohort <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$Cohort))
# batch / chip
hl.merged$batch <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$Batch))
# 3' chemistry
hl.merged$chemistry <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$X3_prime_assay))
# biology
hl.merged$biology <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$Biology))
# pathology
hl.merged$pathology <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$Pathology))
# EBV status
hl.merged$ebv <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$EBVStatus))
# gender
hl.merged$sex <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$Sex))
# age
hl.merged$age <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$Age))
# stage
hl.merged$stage <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$Stage))
# early relapse status
hl.merged$earlyRelapse <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$EarlyRelapse))
# refractory status
hl.merged$refractory <- plyr::mapvalues(x = hl.merged.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$Refractory))

# Visualization transcript information by sample
pdf("~/Documents/sc_HL_relapse/seurat/figures/metrics/VlnPlot-nFeatures_and_pct_MT-by_chemistry.pdf", 
    width = 6, height = 6)
VlnPlot(hl.merged, cols = c("royalblue", "indianred3"),
        features = c("nFeature_RNA", "percent.mt"), group.by = "chemistry",
        sort = "decreasing", log = FALSE,
        ncol = 2, pt.size = 0)
dev.off()
pdf("~/Documents/sc_HL_relapse/seurat/figures/metrics/VlnPlot-nFeatures_and_pct_MT-all_samples-by_chemistry.pdf",
    width = 12, height = 6)
VlnPlot(hl.merged, cols = c("royalblue", "indianred3"),
        features = c("nFeature_RNA", "percent.mt"), split.by = "chemistry",
        log = FALSE,
        ncol = 1, pt.size = 0)
dev.off()

# Create a single vlnplot to trigger the legend
# VlnPlot(hl.merged, cols = c("royalblue", "indianred3"),
#         features = c("percent.mt"), split.by = "chemistry",
#         log = FALSE,
#         ncol = 1, pt.size = 0)

# split cases by chemistry
hl.v2 <- subset(hl.merged, subset = chemistry == "V2") # 25299 genes, 128423 cells
hl.v3 <- subset(hl.merged, subset = chemistry == "V3") # 25299 genes, 69454 cells

# apply filters based on V2 chemistry
hl.v2 <- subset(hl.v2, subset = nFeature_RNA > 200 & nFeature_RNA < 2000 & percent.mt < 10) # 114,544 cells / 28 samples
# apply filters based on V3 chemistry
hl.v3 <- subset(hl.v3, subset = nFeature_RNA > 200 & nFeature_RNA < 3000 & percent.mt < 20) # 51,650 cells / 17 samples

# merge back together
hl.merged <- merge(x = hl.v2, y = hl.v3)

# merge back together
saveRDS(hl.merged, file = "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.merged.rds")
```

## Perform integration

```{r}
# split cases by batch
hl.split <- SplitObject(hl.merged, split.by = "batch")

plan("multiprocess", workers=1)
hl.split <- lapply(X = hl.split, FUN = function(x) {
  x <- SCTransform(x, vars.to.regress = "percent.mt")
})

saveRDS(hl.split, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.split.sctransform.rds")

# !!! Remaining steps must be performed on the GSC projects space (insufficient memory on local computer)
hl.split <- readRDS("hl.split.sctransform.rds")

# select features for downstream integration (NB: 3000 features may be too many for the IntegrateData step)
hl.features <- SelectIntegrationFeatures(object.list = hl.split, nfeatures = 3000)

# implement `future` framework for parallelization
# plan(multisession) # allow parallelization on maximum cores (32 on gphost) - results in error
options(future.globals.maxSize = 15728640000) # allow 15G mem (15*1000*1024^2)
hl.split <- PrepSCTIntegration(object.list = hl.split, anchor.features = hl.features)

# Specify reference - one batch per cohort (choose batch with highest cell count)
#   which(names(hl.split) == "CHIP008") ==> 2
#   which(names(hl.split) == "CHIP020") ==> 10 (has the most cells of all datasets)
#   which(names(hl.split) == "CHIP013") ==> 1
# Note: this dataset is very large and requires a reference for integration (otherwise leads to large-matrix errors)
# reference.dataset <- c(2, 10, 1)
reference.dataset <- which(names(hl.split) == "CHIP020")

# Previously, used 2000 features (default) with log normalization (default, not SCT) and dims=1:20 (default is 1:30).
hl_all.anchors <- FindIntegrationAnchors(object.list = hl.split, 
                                         anchor.features = hl.features, 
                                         reference = reference.dataset,
                                         normalization.method = "SCT")

# without a reference
hl_all.anchors <- FindIntegrationAnchors(object.list = hl.split, 
                                         anchor.features = hl.features, 
                                         normalization.method = "SCT")

# saveRDS(hl_all.anchors, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl_all.anchors.rds")
# options(future.globals.maxSize = 16777216000) # allow 16G mem (16*1000*1024^2)
hl.integrated <- IntegrateData(anchorset = hl_all.anchors, normalization.method = "SCT")
# saveRDS(hl.integrated, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.rds")

# check order of integration - these don't work when using a reference
Tool(object = hl.integrated, slot = "Integration")@sample.tree
hl.integrated@tools$Integration@sample.tree
```

## double check that sample.tree is null using pbmc_small from SeuratData

```{r}
InstallData("panc8")
data("panc8")

pbmc.list <- SplitObject(panc8, split.by = "tech")

pbmc.list <- lapply(X = pbmc.list, FUN = function(x) {
  x <- SCTransform(x)
})

pbmc.features <- SelectIntegrationFeatures(object.list = pbmc.list, nfeatures = 3000)
pbmc.list <- PrepSCTIntegration(object.list = pbmc.list, anchor.features = pbmc.features)
# largest dataset belongs to "smartseq2" which is dataset 3
pbmc.anchors <- FindIntegrationAnchors(object.list = pbmc.list, 
                                         anchor.features = pbmc.features, 
                                         reference = c(3),
                                         normalization.method = "SCT")
pbmc.combined <- IntegrateData(anchorset = pbmc.anchors, normalization.method = "SCT")

# check order of integration
Tool(object = pbmc.combined, slot = "Integration")@sample.tree
pbmc.combined@tools$Integration@sample.tree

## Code by Seurat Lab used to check sample.tree (only works when no reference is used)

library(Seurat)
library(SeuratData)

data("panc8")

obj.list <- SplitObject(panc8, 'tech')
obj.list <- obj.list[1:2]
obj.list <- sapply(obj.list, SCTransform)
features <- SelectIntegrationFeatures(obj.list)
obj.list <- PrepSCTIntegration(obj.list, anchor.features = features)
anchors <- FindIntegrationAnchors(obj.list, anchor.features = features, reference = c(1), normalization.method = "SCT")
integrated <- IntegrateData(anchors)
integrated@tools$Integration@sample.tree
```

## Basic assessment of integration

```{r}
# read back in from the GSC
hl_all.combined <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.rds")

# map metadata
hl_all.combined.idents <- as.data.frame(hl_all.combined$orig.ident)
hl_all.combined.names <- as.character(hl_all.combined.idents$`hl_all.combined$orig.ident`)

# case type
hl_all.combined$case.type <- plyr::mapvalues(x = hl_all.combined.names, 
    from = as.character(final.cohort$Sample), to = as.character(final.cohort$case.type))

# run the standard workflow for visualization and clustering
# hl_all.combined <- ScaleData(hl_all.combined) # this step is not recommended when using scTransform
hl_all.combined <- RunPCA(hl_all.combined)
# get an idea of how many PCs to use --> not a sharp elbow, but should keep at least 20
ElbowPlot(hl_all.combined, ndims = 50)

# UMAP and clustering
hl_all.combined <- RunUMAP(hl_all.combined, reduction = "pca", dims = 1:30)

# see how well the integration went
setwd("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/")
pdf("umap-integrated-by_chip_and_chemistry.pdf", width = 12, height = 5)
DimPlot(hl_all.combined, group.by = c("batch", "chemistry"))
dev.off()

# normalize data for visualization purposes
DefaultAssay(hl_all.combined) <- "RNA"
hl_all.combined <- NormalizeData(hl_all.combined)

# check if JUND is still different between chemistries
p <- VlnPlot(hl_all.combined, features = c("JUND"), 
        split.by = "case.type", 
        pt.size = 0, ncol = 1, log = FALSE, sort = "increasing", 
        assay = "RNA", slot = "data") + xlab("")

p2 <- VlnPlot(hl_all.combined, features = c("JUND"), 
        split.by = "case.type", 
        pt.size = 0, ncol = 1, log = FALSE, sort = "increasing", 
        assay = "integrated", slot = "data") + xlab("")

DefaultAssay(hl_all.combined) <- "integrated"
hl_all.combined <- FindNeighbors(hl_all.combined, reduction = "pca", dims = 1:30)
hl_all.combined <- FindClusters(hl_all.combined, resolution = 0.5)
saveRDS(hl_all.combined, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.clustered.rds")

# Visualization

# split by cohort (diag vs. relapse vs. RLN)
p1 <- DimPlot(hl_all.combined, 
              reduction = "umap", 
              cols = c("relapse"="lightcoral", 
                       "diagnostic"="cornflowerblue",
                       "RLN"="gray40"),
              group.by = "case.type")

# split by cluster
p2 <- DimPlot(hl_all.combined, reduction = "umap", label = TRUE) 
# p2 <- DimPlot(hl_all.combined, reduction = "umap", group.by = "celltype", label = TRUE) 

library(ggplot2)
library(gridExtra)
library(cowplot)
plot_grid(p1, p2, align = "h", ncol = 2, rel_widths = c(3/7, 4/7))

# Plot clusters by themselves
pdf("umap-clusters.pdf", width = 8, height = 7)
p2
dev.off()

# Plot cohort-level and cluster-level umap
pdf("umap-integrated-by_cohort_and_cluster.pdf", width = 12, height = 5)
grid.arrange(p1, p2, ncol=2)
dev.off()

# Visualize diagnostic and relapse cases separately
pdf("umap-per_cohort-clusters.pdf", width = 13, height = 5)
DimPlot(hl_all.combined, reduction = "umap", split.by = "case.type")
dev.off()

# get list of highly-variable genes
hl.all.hvg <- hl_all.combined@assays$integrated@var.features
write.table(hl.all.hvg, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.hgv.txt", sep = "\t", quote = FALSE)
```

## Create heatmap of average expression of clusters across component markers

```{r}
#hl_all.combined <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.clustered.rds")
# create heatmap split by component - to add PRF1, CD11B (ITGAM) [[CTL]], CXCR3, CD1D, and IL35 (IL12A) [[B cell]]
component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", sep = "\t", header = TRUE, fill = TRUE)

# component markers that are in list of HVG:
markers.to.plot <- component.mappings$Gene

# only look at markers that are in the HVG list
markers.to.plot <- subset(hl.all.hvg, hl.all.hvg %in% markers.to.plot)

cluster.averages <- AverageExpression(hl_all.combined, 
                                      features = markers.to.plot,
                                      return.seurat = TRUE)
hl.integrated.cluster.avgs <- GetAssayData(object = cluster.averages, slot = "scale.data")
hl.integrated.components.cluster.avgs <- hl.integrated.cluster.avgs[ rownames(hl.integrated.cluster.avgs) %in% markers.to.plot , ]

# remove genes that are not part of HVG list (e.g. IL4, IDO1, SDC1, NRP1)
component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
rownames(component.mappings) <- component.mappings$Gene

# order by component then gene name so that the heatmap legend makes sense
components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
component.mappings$Gene <- NULL
component.mappings$Alt_name <- NULL

# reorder expression matrix to match component mappings
hl.integrated.components.cluster.avgs.ordered <- hl.integrated.components.cluster.avgs[ match(rownames(components.ordered), rownames(hl.integrated.components.cluster.avgs)), ]

pdf("heatmap-component_markers-avg_expr_per_cluster-tall.pdf", width = 12, height = 15)
p4 <- pheatmap(hl.integrated.components.cluster.avgs.ordered, 
         annotation_row = component.mappings,
         cluster_rows = FALSE, 
         annotation_names_row = FALSE, 
         angle_col = 45, 
         gaps_row = c(13,19,21,23,26,31,33,37,38,39,40,43,50,52), border_color = NA)

pdf("heatmap-component_markers-avg_expr_per_cluster.pdf", width = 13, height = 5)
p4.rotated <- pheatmap(t(hl.integrated.components.cluster.avgs.ordered), 
         annotation_col = component.mappings,
         cluster_cols = FALSE, 
         annotation_names_col = FALSE, 
         gaps_col = c(13,19,21,23,26,31,33,37,38,39,40,43,50,52), 
         angle_col = 45, 
         border_color = NA)
dev.off()

```

## Produce single-cell level heatmaps of expression for EACH cluster

```{r}
NUM_CLUSTERS = 26

hl.integrated.data <- hl_all.combined@assays$SCT@data
# this is a 1559 (features) X 76607 (barcodes) matrix

# retrieve expression for each cell across the component genes of interest
hl.integrated.markers <- hl.integrated.data[ rownames(hl.integrated.data) %in% markers.to.plot , ]

# create heatmap split by component
component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", sep = "\t", header = TRUE, fill = TRUE)
# remove genes that are not part of HVG list (IL4, IDO1, NRP1)
component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
# order by component then gene name so that the heatmap legend makes sense
rownames(component.mappings) <- component.mappings$Gene
components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
components.ordered$Gene <- NULL
components.ordered$Alt_name <- NULL

# reorder expression matrix to match component mappings
hl.integrated.markers.ordered <- hl.integrated.markers[ row.names(components.ordered), ]

# to create separate expr matrices for each cluster, need mappings for cell --> cluster
hl_integrated.clusters <- hl_all.combined$seurat_clusters
# convert to char
clusters.char <- levels(hl_integrated.clusters)[hl_integrated.clusters] # ordered vector of cluster ids (0 through 26)
barcodes.cluster <- attributes(hl_integrated.clusters)$names # ordered vector of barcodes (76,607 unique entries)
clusters.df <- data.frame(barcodes.cluster, clusters.char)

# extract barcodes corresponding to clusters (an array of vectors?)
# pseudocode: 

expr.markers <- vector('list', NUM_CLUSTERS)
heatmaps <- vector('list', NUM_CLUSTERS)
hl.seurat <- vector('list', NUM_CLUSTERS)

for (c in 1:NUM_CLUSTERS) {
  hl.seurat[[c]] <- subset(hl_all.combined, idents = c-1) # filter by cluster
  hl.seurat[[c]] <- subset(hl.seurat[[c]], features = markers.to.plot) # filter by genes
  
  # retrieve expression for each cell across the component genes of interest
  expr.temp <- hl.seurat[[c]]@assays$SCT@data
  expr.markers[[c]] <- expr.temp[ rownames(expr.temp) %in% markers.to.plot , ]
  
  # reorder expression matrix to match component mappings
  expr.markers[[c]] <- expr.markers[[c]][ row.names(components.ordered), ]

  filename <- paste("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/heatmaps/heatmap-cluster", 
                    c-1, ".pdf", sep = "")
  pdf(filename, width = 8, height = 8)
  heatmaps[[c]] <- pheatmap(expr.markers[[c]], 
         annotation_row = components.ordered, 
         cluster_rows = FALSE, 
         annotation_names_row = FALSE, 
         angle_col = 45, 
         gaps_row = c(13,19,21,23,26,31,33,37,38,39,40,43,50,52), 
         border_color = NA, show_colnames = FALSE, scale = "none")
  dev.off()
}

```

## Rename clusters based on feature plots

```{r}
# Tomo's latest mappings (as of November 19, 2020)
hl_all.combined <- RenameIdents(hl_all.combined,
                                `0` = "Naive B cell",
                                `1` = "Naive T cell",
                                `2` = "Naive B cell",
                                `3` = "Naive T cell",
                                `4` = "T reg",
                                `5` = "Naive B cell",
                                `6` = "T helper",
                                `7` = "CTL",
                                `8` = "T helper",
                                `9` = "Naive B cell",
                                `10` = "T reg/T helper",
                                `11` = "B cell",
                                `12` = "Naive CD8 cell",
                                `13` = "Naive B cell",
                                `14` = "Naive T cell",
                                `15` = "T reg",
                                `16` = "Naive B cell",
                                `17` = "Naive T cell",
                                `18` = "B cell",
                                `19` = "NK cell",
                                `20` = "GCB and T cell proliferation",
                                `21` = "Naive B cell",
                                `22` = "Naive B cell",
                                `23` = "Macrophage/pDC",
                                `24` = "Plasma cell",
                                `25` = "Naive B cell")
hl_all.combined$celltype <- Idents(hl_all.combined)
saveRDS(hl_all.combined, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.annotated.rds")

# Now plot the new annotations
pdf("umap-annotated_clusters.pdf", width = 8, height = 7)
DimPlot(hl_all.combined, reduction = "umap", label = TRUE, pt.size = 0.2)
dev.off()

# create a barplot to show the numbers of cells in each cluster / component type
celltype.numcells <- as.data.frame(table(hl_all.combined$celltype))
celltype.numcells$temp <- "NA"
colnames(celltype.numcells) <- c("celltype", "numcells", "temporarycol")

p <- ggplot(celltype.numcells, aes(x = reorder(celltype, -numcells), y = numcells, fill=temporarycol)) + 
      geom_bar(stat = "identity", width = 0.8) + 
      scale_fill_manual(values=c("darkblue"))+
      xlab("") + ylab("") + 
      ylim(0, 78000) +
      theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
      theme(panel.spacing = unit(1, "lines")) +
      geom_text(aes(label=numcells), hjust=0.5, vjust=-1, size=3, color="gray47") +
      theme(legend.position="right")

# create barplot showing proportion of cells per cell type based on case type
celltype.numcells2 <- as.data.frame(table(hl_all.combined$celltype, hl_all.combined$case.type))
colnames(celltype.numcells2) <- c("celltype", "case_type", "numcells")

p.log <- ggplot(celltype.numcells2, aes(x = reorder(celltype, -numcells), y = numcells, fill = case_type)) + 
      geom_bar(position = "fill", stat = "identity", width = 0.8) + 
      scale_y_continuous(labels = scales::percent_format()) +
      xlab("") + ylab("") + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
      theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
      theme(panel.spacing = unit(1, "lines")) + 
      theme(legend.title = element_blank()) +
      theme(legend.position="right")

# create a similar barplot, but showing proportion of cells per cell type that are V2 vs. V3
celltype.numcells3 <- as.data.frame(table(hl_all.combined$celltype, hl_all.combined$chemistry))
colnames(celltype.numcells3) <- c("celltype", "chemistry", "numcells")

p.chem <- ggplot(celltype.numcells3, aes(x = reorder(celltype, -numcells), y = numcells, fill = chemistry)) + 
      geom_bar(position = "fill", stat = "identity", width = 0.8) + 
      scale_y_continuous(labels = scales::percent_format()) +
      xlab("") + ylab("") + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
      theme(panel.spacing = unit(1, "lines")) + 
      theme(legend.title = element_blank()) +
      theme(legend.position="right") +
      scale_fill_manual(values=c("cornflowerblue", "indianred3"), labels=c("V2 assay ", "V3 assay "))

pdf("barplot-num_cells_and_cohort_ratios_per_cluster.pdf", width = 8, height = 8)
grid.arrange(p, p.log, p.chem, ncol=1, nrow=3, widths=c(8), heights=c(4,3,6))
dev.off()

# create a stacked barplot to show the proportion of each cell identity by sample
sample.celltype_num <- as.data.frame(table(hl_all.combined$orig.ident, hl_all.combined$celltype))
colnames(sample.celltype_num) <- c("sample_id", "cell_identity", "num_cells")

# order by proportion of naive B cells
sample.celltype_num <- sample.celltype_num %>% 
  dplyr::group_by(sample_id) %>% dplyr::mutate(total_cells = sum(as.numeric(num_cells)))
sample.celltype_num <- as.data.frame(sample.celltype_num %>% dplyr::mutate(proportion = as.numeric(num_cells) / total_cells))
# get order of samples by focusing only on proportion of naive B cells
temp <- subset(sample.celltype_num, 
               sample.celltype_num$cell_identity == "Naive B cell")
temp <- temp[ order(temp$cell_identity, -temp$proportion) , ]
# set the order in the full dataframe
sample.celltype_num$sample_id <- factor(sample.celltype_num$sample_id, levels = temp$sample_id)

pdf("barplot-cell_types_per_sample.pdf", width = 9, height = 4.5)
ggplot(sample.celltype_num, 
            aes(x=sample_id, y=num_cells, fill=cell_identity)) +
  geom_bar(position = "fill", stat = "identity")+
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5)) +
  scale_fill_manual(name = "Cell Identity",
                    values = c("Naive B cell" = "darkslateblue",
                               "Naive T cell" = "slategray4",
                               "T reg" = "aquamarine3", #aquamarine3
                               "T helper" = "hotpink3",
                               "CTL" = "olivedrab4",
                               "T reg/T helper" = "bisque3",#bisque3
                               "B cell"="royalblue1",#royalblue1
                               "Naive CD8 cell"="firebrick3",#firebrick3
                               "NK cell"="darkorange",#darkorange
                               "GCB and T cell proliferation"="mediumpurple1",#mediumpurple1
                               "Macrophage/pDC"="green3",
                               "Plasma cell"="magenta3"#magenta3
                               )) +

  xlab("") +
  ylab("")
dev.off()

```

## Assign cell types using celldex

See https://bioconductor.org/packages/release/data/experiment/vignettes/celldex/inst/doc/userguide.html

```{r}
library(celldex)
library(SingleR)
library(SingleCellExperiment)

# Subset 1: focus on macrophage / pDC cluster
hl.macrophage_pDC <- subset(x = hl_all.combined, subset = celltype == "Macrophage/pDC")
# Result: 514 cells (from a total of 166194; 0.3% of cells are kept)

# Subset 2: A larger population of cells including the macrophage / pDC cluster
clusters <- c("Macrophage/pDC", "Plasma cell", "NK cell", "CTL", "GCB and T cell proliferation")
hl.subset <- subset(x = hl_all.combined, subset = celltype %in% clusters)
# Result: 13711 cells --> 8% of cells are kept

# pull matrix from Seurat Object
singler_data <- as.SingleCellExperiment(hl.macrophage_pDC)

# get reference data
ref <- HumanPrimaryCellAtlasData() # deprecated
ref <- BlueprintEncodeData()
# ref <- ImmGenData()
# ref <- DatabaseImmuneCellExpressionData()
# ref <- NovershternHematopoieticData()
# ref <- MonacoImmuneData()

# Get cell type information
cell.main <- SingleR(singler_data, ref, labels = ref$label.main)
cell.fine <- SingleR(singler_data, ref, labels = ref$label.fine)

# Store cell type information into Seurat Object
# hl.macrophage_pDC[['HPCA.labels.main']] <- cell.main$labels
hl.macrophage_pDC[['Blueprint.labels.main']] <- cell.main$labels
hl.macrophage_pDC[['Blueprint.labels.fine']] <- cell.fine$labels
# hl.macrophage_pDC[['ImmGen.labels.main']] <- cell.main$labels
# hl.macrophage_pDC[['DbImmuneCell.labels.main']] <- cell.main$labels
# hl.macrophage_pDC[['Novershtern.labels.main']] <- cell.main$labels
# hl.macrophage_pDC[['Monaco.labels.main']] <- cell.main$labels

# hl.macrophage_pDC <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.macrophage_pDC.celldex.rds")
hl.blueprint <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.blueprint.rds")
cell.main.BP <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/cell.main.Blueprint.integrated.rds")
cell.fine.BP <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/cell.fine.Blueprint.integrated.rds")

# visualize scores
plotScoreHeatmap(cell.main)
plotDeltaDistribution(cell.main, ncol = 3)

# visualize distribution of cells across labels and clusters
tab <- table()
tab <- table(Assigned = cell.main$pruned.labels, 
             Cluster = colLabels(hl.macrophage_pDC))

# Adding a pseudo-count of 10 to avoid strong color jumps with just 1 cell.
library(pheatmap)
pheatmap(log2(tab+10), color=colorRampPalette(c("white", "blue"))(101))

# visualize on umap and with heatmaps
p1 <- DimPlot(hl.macrophage_pDC, reduction = "umap", group.by = "HPCA.labels.main") + 
  labs(color = "HPCA") 
p2 <- DimPlot(hl.macrophage_pDC, reduction = "umap", group.by = "Blueprint.labels.main") +
  labs(color = "Blueprint Encode") 
p3 <- DimPlot(hl.macrophage_pDC, reduction = "umap", group.by = "ImmGen.labels.main") +
  labs(color = "ImmGen") 
p4 <- DimPlot(hl.macrophage_pDC, reduction = "umap", group.by = "DbImmuneCell.labels.main") +
  labs(color = "DB Immune Cell") 
p5 <- DimPlot(hl.macrophage_pDC, reduction = "umap", group.by = "Novershtern.labels.main") +
  labs(color = "Novershtern") 
p6 <- DimPlot(hl.macrophage_pDC, reduction = "umap", group.by = "Monaco.labels.main") +
  labs(color = "Monaco") 
pdf("umap-macrophage_pDC_celldex.pdf", width = 14, height = 7)
grid.arrange(p1, p2, p3, p4, p5, p6, ncol=3)
dev.off()

# subset
DimPlot(hl.subset, reduction = "umap", group.by = "Blueprint.labels.main") +
  labs(color = "Blueprint Encode") 

# full dataset
hl.blueprint[['Blueprint.labels.main']] <- cell.main.BP$labels
hl.blueprint[['Blueprint.labels.fine']] <- cell.fine.BP$labels
p1 <- DimPlot(hl.blueprint, reduction = "umap", group.by = "Blueprint.labels.main") +
  labs(color = "Blueprint: main labels") 
p2 <- DimPlot(hl.blueprint, reduction = "umap", group.by = "Blueprint.labels.fine") +
  labs(color = "Blueprint: fine labels") 

```

## Draw heatmaps for celldex annotations

```{r}
# Idents(hl.macrophage_pDC) <- hl.macrophage_pDC$HPCA.labels.main
# Idents(hl.macrophage_pDC) <- hl.macrophage_pDC$Blueprint.labels.main
# Idents(hl.macrophage_pDC) <- hl.macrophage_pDC$ImmGen.labels.main
# Idents(hl.macrophage_pDC) <- hl.macrophage_pDC$DbImmuneCell.labels.main
# Idents(hl.macrophage_pDC) <- hl.macrophage_pDC$Novershtern.labels.main
Idents(hl.macrophage_pDC) <- hl.macrophage_pDC$Monaco.labels.main
```

```{r}
# create heatmap split by component
component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", sep = "\t", header = TRUE, fill = TRUE)

# component markers that are in list of HVG:
markers.to.plot <- component.mappings$Gene

# only look at markers that are in the HVG list
markers.to.plot <- subset(hl.all.hvg, hl.all.hvg %in% markers.to.plot)

cluster.averages <- AverageExpression(hl.macrophage_pDC, 
                                      features = markers.to.plot,
                                      return.seurat = TRUE)
hl.macrophage_pDC.cluster.avgs <- GetAssayData(object = cluster.averages, slot = "scale.data")
hl.macrophage_pDC.components.cluster.avgs <- hl.macrophage_pDC.cluster.avgs[ rownames(hl.macrophage_pDC.cluster.avgs) %in% markers.to.plot , ]

# remove genes that are not part of HVG list (e.g. IL4, IDO1, SDC1, NRP1)
component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
rownames(component.mappings) <- component.mappings$Gene

# order by component then gene name so that the heatmap legend makes sense
components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
component.mappings$Gene <- NULL
component.mappings$Alt_name <- NULL

# reorder expression matrix to match component mappings
hl.macrophage_pDC.components.cluster.avgs.ordered <- hl.macrophage_pDC.components.cluster.avgs[ match(rownames(components.ordered), rownames(hl.macrophage_pDC.components.cluster.avgs)), ]
```

```{r}
# ref_name <- "HPCA" # 6 x 10
# ref_name <- "Blueprint" # 4.75 x 10
# ref_name <- "ImmGen" # 5 X 10
# ref_name <- "DBImmuneCell" # 4.75 x 10
# ref_name <- "Novershtern" # 5.5 x 10
ref_name <- "Monaco" # 5 x 10

pdf(paste("heatmap-macrophage_pDC-", ref_name, "-component_markers-avg_expr_per_celltype.pdf", sep = ""), 
    width = 5, height = 10)
pheatmap(hl.macrophage_pDC.components.cluster.avgs.ordered, 
         annotation_row = component.mappings,
         cluster_rows = FALSE, 
         annotation_names_row = FALSE, 
         angle_col = 90,
         gaps_row = c(11,17,19,22,27,29,33,34,35,36,39,46,48), 
         border_color = NA, main = ref_name)
dev.off()

```

## Isolate CD4+ CTL-cells

```{r}
hl_all.combined <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.clustered.rds")
# hl_all.combined <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.annotated.rds")
# subset based on:
#   Expression of CD3, CD4
#   No expression of CD56, CD8, CD19, and CD68

# First get CTL cluster
hl.ctl <- subset(hl_all.combined, subset = celltype == "")

# Subset 1: focus on macrophage / pDC cluster
# hl.macrophage_pDC <- subset(x = hl_all.combined, subset = celltype == "Macrophage/pDC")
# Result: 514 cells (from a total of 166194; 0.3% of cells are kept)
```

## Find conserved markers (to help refine cluster annotations)

```{r}
library(metap)
library(MAST)
library(dplyr)

# change default identities to cell identities (cell types)
Idents(hl_all.combined) <- hl_all.combined$celltype

## start with one cluster to test (then do a for loop)
# naiveBcell.markers <- FindConservedMarkers(hl_all.combined, ident.1 = "Naïve B cells", grouping.var = "case.type")

# iterate through all the cell identities
cell.identities <- as.vector(unique(Idents(hl_all.combined)))
# store the results in a vector of data frames
# initialize empty vector of results
markers.results <- vector('list', length(cell.identities))

# Approach 1: find markers that are conserved between two groups (cell identity vs. all other identities)
for (i in 1:length(cell.identities)) {
  markers.results[[i]] <- FindConservedMarkers(hl_all.combined, 
                                               ident.1 = cell.identities[i],  
                                               grouping.var = "case.type")
}
# Top (up to) 3 *conserved* markers for each cluster / cell type / cell identity:
#1. Naive T cells - CCR7, IL7R
#2. Tregs - CTLA4, IL32, TNFRSF18
#3. Helper T cells - TRAC, CD2
#4. Macrophages - CLEC10A, CSTA
#5. B cells - IGHM, HLA-DRA, MS4A1, CD74
#6. CTLs - CCL5, GZMK, NKG7
#7. GCBs - RGS13, CCDC144A
#8. Plasmocytoid DCs - GZMB, PTGDS
#9. NK cells - GNLY, NKG7
#10. B and T mix - no conserved markers
#11. Plasma cells - DERL3
#12. B cells, macrophages - RSAD2, IFIT3
#13. Treg proliferation - MKI67, BIRC5

# Plot top conserved marker using FeaturePlots
top.conserved.markers <- c("DERL3", "MKI67", "BIRC5")
FeaturePlot(hl_all.combined, features = top.conserved.markers, min.cutoff = "q9")

# plot in groups
# Group 1: T cells
t.markers <- c("IL7R", "IL32", "TRAC", "CCL5")
FeaturePlot(hl_all.combined, features = t.markers, min.cutoff = "q9")

# Group 2: B cells
b.markers <- c("MS4A1", "HLA-DRA", "IGHM", "CD74")
FeaturePlot(hl_all.combined, features = b.markers, min.cutoff = "q9")

# Group 3: all other cells
other.markers <- c("CSTA", # macrophages
                   "RGS13", # GCBs
                   "PTGDS", # pDCs
                   "GNLY", # NK cells
                   "MKI67", # Treg proliferation
                   "DERL3" # plasma cells
                   )
FeaturePlot(hl_all.combined, features = other.markers, min.cutoff = "q9")

# APPROACH 2: use FindAllMarkers (finds markers that are differentially expressed in each identity group vs all others)

markers.results2 <- FindAllMarkers(hl_all.combined, assay = "RNA", test.use = "MAST")
# filter for adjusted p-value < 0.05
markers.results2 <- filter(markers.results2, markers.results2$p_val_adj < 0.05)

write.table(markers.results2, "~/Documents/sc_HL_relapse/seurat/output/LATEST/hl_all.combined-all_markers.txt", sep = "\t", row.names = FALSE, quote = FALSE)

# get top 3 markers for each cluster
markers.results2.top3 <- do.call(rbind, 
                lapply(split(markers.results2, markers.results2$cluster), function(x) x[order(-x$avg_logFC)[1:3],]))
markers.results2.top1 <- do.call(rbind, 
                lapply(split(markers.results2, markers.results2$cluster), function(x) x[order(-x$avg_logFC)[1],]))

# Top markers for each cluster (based on differential expression to all other clusters)
top.conserved.markers <- markers.results2.top1$gene
FeaturePlot(hl_all.combined, features = top.conserved.markers, min.cutoff = "q9")

# Visualize conserved cell type markers:
# conserved.markers <- c(t.markers, b.markers, other.markers)
dp <- DotPlot(
    object = hl_all.combined,
    features = top.conserved.markers, 
    dot.scale = 8,
    split.by = "case.type", cols = c("orange", "blue", "red")) + 
    RotatedAxis() + xlab("") + ylab("") + coord_flip()

```

## Differential expression analysis - testing

Check out https://nbisweden.github.io/excelerate-scRNAseq/session-de/session-de-methods.html for a review of DE methods that can be applied to scRNAseq data (not just those integrated in Seurat)

```{r}
# change default assay to RNA
DefaultAssay(hl_all.combined) <- "RNA"
# normalize and scale data for DE testing
hl_all.combined <- NormalizeData(hl_all.combined)
hl_all.combined <- ScaleData(hl_all.combined)

hl_all.combined$celltype.case <- paste(Idents(hl_all.combined), hl_all.combined$case.type, sep = "_")
hl_all.combined$celltype <- Idents(hl_all.combined) # keep a copy of the cluster annotations
Idents(hl_all.combined) <- hl_all.combined$case.type # make case type the main identity (for DE analysis)

## try out all the differential expression methods:
# de.methods <- c("wilcox", # NA's for p-values --> generic method
#                 "bimod", # low discriminatory power with p-value (mostly 0's)
#                 "roc", # 0 or 1 are good; 0.5 is bad classifer
#                 "t", # 9 / 54 genes have p-value > 0 (and < 0.05)
#                 "LR", #  9 / 54 genes have p-value > 0 (and < 0.05)
#                 "MAST", #  2 / 54 genes have p-value > 0 (and < 0.05) --> **only scRNAseq method
#                 "DESeq2", # doesn't work --> designed for bulk RNAseq
#                 "negbinom", # doesn't work --> generic method
#                 "poisson" # doesn't work --> generic method
# )

# initialize empty vector of results
# de.results <- vector('list', length(de.methods))

# find DE genes between all diagnostic vs. all relapse cases (not cluster-specific) for each method
# for (i in 1:length(de.methods)) {
  # de.results[[i]] <- FindMarkers(hl.both.combined, ident.1 = "relapse", ident.2 = "diagnostic", 
                          # test.use = de.methods[i],
                          # min.pct = 0.25, logfc.threshold = 0.25)
# }
```


For FindMarkers function, NB: group 1 vs. group 2 indicates for avg_logFC (log fold-chage of the average expression between the two groups), a POSITIVE value indicates the gene is MORE HIGHLY EXPRESSED in the FIRST group (i.e. group 1).

*Important:* Use the uncorrected raw (but normalized) expression data to perform DE (set default assay to RNA)

## Functions to perform DE comparisons at the cluster (cell type) level

```{r}
# function to find DE genes between all pairs

get_de_genes <- function(comparators, de.test, cell_counts.df) {
  
  # get list of comparators that have fewer than 100 cells
  cell_counts.df <- cell_counts.df[ cell_counts.df$Var1 %in% comparators, ] # only look at comparators of interest
  low_cell_counts.comparators <- cell_counts.df[ cell_counts.df$Freq < 3, ]$Var1
  
  NUM_CLUSTERS = length(comparators) / 2
  # de_results <- vector('list', NUM_CLUSTERS)
  de_results <- list()

  for (i in 1:NUM_CLUSTERS) {
    index.grp1 <- (i*2)-1
    index.grp2 <- i*2
    
    # check that both grp1 and grp2 are NOT in the low cell counts list
    if ( !(comparators[index.grp1] %in% low_cell_counts.comparators) && 
         !(comparators[index.grp2] %in% low_cell_counts.comparators)) {
      
        # slot = "data" (normalized counts)
        tempResults <- FindMarkers(hl_all.combined, 
                                   assay = "RNA", slot = "data",
                                   ident.1 = comparators[index.grp1], 
                                   ident.2 = comparators[index.grp2], 
                                   test.use = de.test, 
                                   latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
                                   min.pct = 0.25, 
                                   logfc.threshold = 0)
        # make gene a column
        tempResults$gene <- rownames(tempResults)
        # add comparison description as a column
        tempResults$comparison <- paste(comparators[index.grp1], "_VS_", comparators[index.grp2], sep = "")
        rownames(tempResults) <- NULL
        # change order of columns
        cols <- c("comparison", "gene", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")
        tempResults <- tempResults[ cols ]
        # sort by descending fold change
        tempResults <- tempResults[order(-tempResults$avg_log2FC), ]
        # filter out non-interesting genes (i.e. *orf*, ^RP, and ^MT- genes)
        tempResults <- tempResults[ !(grepl("(^RP)|(MT-)|(*orf*)", tempResults$gene)), ]
        # remove results that have an adjusted p-value of 1
        tempResults <- dplyr::filter(tempResults, tempResults$p_val_adj < 1)
        
        # !!! NB: this step will fail if tempResults is empty (e.g. all adjusted p-values are 1)
        # store the processed results in the final result array
        if (nrow(tempResults) > 0) {
          de_results[[unique(tempResults$comparison)]] <- tempResults  
        }
      }
  }
  return(de_results)
}

# function to merge results into single dataframe
merge_de_results <- function(results.vector) {
  # initialize dataframe
  merged.df <- data.frame(comparison = character(),
                          gene = character(),
                          p_val = double(),
                          avg_logFC = double(),
                          pct.1 = double(),
                          pct.2 = double(),
                          p_val_adj = double())
  
  NUM_CLUSTERS = length(results.vector)
  for (i in 1:NUM_CLUSTERS) {
    merged.df <- rbind(merged.df, results.vector[[i]])
  }
  return(merged.df)
}
```

## For sanity purposes - test out the different methods: wilcox, bimod, roc, t, negbinom, poisson, LR, MAST, DESeq2

```{r}
de.test.mast <- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "MAST", assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

de.test.wilcox <- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "wilcox", 
                       assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

de.test.bimod <- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "bimod", 
                       assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

de.test.roc <- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "roc", 
                       assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

de.test.t<- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "t", 
                       assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

de.test.negbinom <- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "negbinom", 
                       assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

de.test.poisson <- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "poisson", 
                       assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

de.test.LR <- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "LR", 
                       assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

de.test.DESeq2 <- FindMarkers(hl.both.combined, 
                       ident.1 = "Macrophages_diagnostic", 
                       ident.2 = "Macrophages_relapse", 
                       test.use = "DESeq2", 
                       assay = "RNA", slot = "data",
                       latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

```

## Differential expression analyses for all comparisons, by cell type (annotated clusters) and by cluster  

*Remeber to use raw + normalized data for DE analysis*

```{r}
# change default assay to RNA
DefaultAssay(hl_all.combined) <- "RNA"
# normalize and scale data for DE testing
hl_all.combined <- NormalizeData(hl_all.combined)

# define working directory:
work.dir <- "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/DE_analysis/"
```

```{r}
## #########################################################
## 3' assay V2 vs. V3 chemistry
## #########################################################

Idents(hl_all.combined) <- hl_all.combined$chemistry
comparators <- sort(unique(hl_all.combined$chemistry))

# though unnecessary for this scenario, store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$chemistry))

# perform DE analysis, where odd indices = diagnostic, even indices = relapse
de.results <- FindMarkers(hl_all.combined, 
                           assay = "RNA", slot = "data",
                           ident.1 = comparators[1], 
                           ident.2 = comparators[2], 
                           test.use = "MAST", 
                           latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
                           min.pct = 0.25, 
                           logfc.threshold = 0)
# make gene a column
de.results$gene <- rownames(de.results)
# add comparison description as a column
de.results$comparison <- paste(comparators[1], "_VS_", comparators[2], sep = "")
rownames(de.results) <- NULL
# change order of columns
cols <- c("comparison", "gene", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")
de.results <- de.results[ cols ]
# sort by descending fold change
de.results <- de.results[order(-de.results$avg_log2FC), ]
# filter out non-interesting genes (i.e. *orf*, ^RP, and ^MT- genes)
de.results <- de.results[ !(grepl("(^RP)|(MT-)|(*orf*)", de.results$gene)), ]
# remove results that have an adjusted p-value of 1
de.results <- dplyr::filter(de.results, as.numeric(de.results$p_val_adj) < 1)

write.table(de.results, 
            paste(work.dir, "de.V2_v_V3.txt", sep=""),
            sep = "\t", quote = FALSE, row.names = FALSE)

```

```{r}
## #########################################################
## Cluster-level DE analysis: V2 vs. V3 assay
## #########################################################

## ------ cell-type level -------

# define comparators (for annotated cell types)
hl_all.combined$celltype.chem <- paste(hl_all.combined$celltype, hl_all.combined$chemistry, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$celltype.chem
comparators <- sort(unique(hl_all.combined$celltype.chem))

# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$celltype.chem))

# perform DE analysis, where odd indices = diagnostic, even indices = relapse
de.V2_vs_V3.celltypes <- get_de_genes(comparators = comparators, de.test = "MAST", cell_counts.df = cell_counts.df)
de.V2_vs_V3.celltypes.df <- merge_de_results(de.V2_vs_V3.celltypes)
write.table(de.V2_vs_V3.celltypes.df, 
            paste(work.dir, "de.V2_vs_V3.celltypes.txt", sep=""),
            sep = "\t", quote = FALSE, row.names = FALSE)

## ------ cluster level -------

hl_all.combined$cluster.chem <- paste(hl_all.combined$seurat_clusters, hl_all.combined$chemistry, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$cluster.chem
comparators <- sort(unique(hl_all.combined$cluster.chem))

# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$cluster.chem))

# perform DE analysis, where odd indices = diagnostic, even indices = relapse
de.V2_vs_V3.clusters <- get_de_genes(comparators = comparators, de.test = "MAST", cell_counts.df = cell_counts.df)
de.V2_vs_V3.clusters.df <- merge_de_results(de.V2_vs_V3.clusters)
write.table(de.V2_vs_V3.clusters.df, 
            paste(work.dir, "de.V2_vs_V3.clusters.txt", sep=""),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

```{r}
## #########################################################
## Cluster-level DE analysis: diagnostic vs. relapse
## #########################################################

## ------ cell-type level -------

# define comparators (for annotated cell types)
hl_all.combined$celltype.case <- paste(hl_all.combined$celltype, hl_all.combined$case.type, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$celltype.case
comparators <- sort(unique(hl_all.combined$celltype.case))

# remove RLN cases
comparators <- comparators[ !grepl("_RLN", comparators)]
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$celltype.case))

# perform DE analysis, where odd indices = diagnostic, even indices = relapse
de.diag_vs_rel.celltypes <- get_de_genes(comparators = comparators, de.test = "MAST", cell_counts.df = cell_counts.df)
de.diag_vs_rel.celltypes.df <- merge_de_results(de.diag_vs_rel.celltypes)
write.table(de.diag_vs_rel.celltypes.df, 
            paste(work.dir, "de.diag_vs_rel.cell_types.txt", sep=""),
            sep = "\t", quote = FALSE, row.names = FALSE)

## ------ cluster level -------

hl_all.combined$cluster.case <- paste(hl_all.combined$seurat_clusters, hl_all.combined$case.type, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$cluster.case
comparators <- sort(unique(hl_all.combined$cluster.case))

# remove RLN cases
comparators <- comparators[ !grepl("_RLN", comparators)]
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$cluster.case))

# perform DE analysis, where odd indices = diagnostic, even indices = relapse
de.diag_vs_rel.clusters <- get_de_genes(comparators = comparators, de.test = "MAST", cell_counts.df = cell_counts.df)
de.diag_vs_rel.clusters.df <- merge_de_results(de.diag_vs_rel.clusters)
write.table(de.diag_vs_rel.clusters.df, 
            paste(work.dir, "de.diag_vs_rel.clusters.txt", sep=""),
            sep = "\t", quote = FALSE, row.names = FALSE)


## #########################################################
## Cluster-level DE analysis: diagnostic vs. refractory
## #########################################################

## ------ cell-type level -------

# define comparators
hl_all.combined$celltype.refractory <- paste(hl_all.combined$celltype, hl_all.combined$refractory, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$celltype.refractory 
comparators <- sort(unique(hl_all.combined$celltype.refractory))

# remove comparators associated "_relapseNotRefractory" and RLN (i.e. "_NA") cases
comparators <- comparators[ !grepl("_(relapseNotRefractory|RLN)", comparators)]
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$celltype.refractory))

# perform the DE analysis
de.diag_vs_refractory.celltypes <- get_de_genes(comparators = comparators, de.test = "MAST", cell_counts.df = cell_counts.df)
de.diag_vs_refractory.celltypes.df <- merge_de_results(de.diag_vs_refractory.celltypes)
write.table(de.diag_vs_refractory.celltypes.df, 
            paste(work.dir, "de.diag_vs_refractory.cell_types.txt", sep = ""), 
            sep = "\t", quote = FALSE, row.names = FALSE)

## ------ cluster level -------

hl_all.combined$cluster.refractory <- paste(hl_all.combined$seurat_clusters, hl_all.combined$refractory, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$cluster.refractory 
comparators <- sort(unique(hl_all.combined$cluster.refractory))

# remove comparators associated "_relapseNotRefractory" and RLN (i.e. "_NA") cases
comparators <- comparators[ !grepl("_(relapseNotRefractory|RLN)", comparators)]
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$cluster.refractory))

# perform the DE analysis
de.diag_vs_refractory.clusters <- get_de_genes(comparators = comparators, de.test = "MAST", cell_counts.df = cell_counts.df)
de.diag_vs_refractory.clusters.df <- merge_de_results(de.diag_vs_refractory.clusters)
write.table(de.diag_vs_refractory.clusters.df, 
            paste(work.dir, "de.diag_vs_refractory.clusters.txt", sep = ""), 
            sep = "\t", quote = FALSE, row.names = FALSE)


## #########################################################
## Cluster-level DE analysis: diagnostic vs. early relapse, diagnostic vs. late relapse
## #########################################################

## ------ cell type level -------

# define comparators
hl_all.combined$celltype.earlyrelapse <- paste(hl_all.combined$celltype, hl_all.combined$earlyRelapse, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$celltype.earlyrelapse 
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$celltype.earlyrelapse))

# diagnostic vs. early relapse
comparators <- sort(unique(hl_all.combined$celltype.earlyrelapse))
# remove the "_lateRelapse" and RLN cases
comparators <- comparators[ !grepl("_(lateRelapse|RLN)", comparators)]

# perform the DE analysis
de.diag_vs_early_relapse.celltypes <- get_de_genes(comparators, de.test = "MAST", cell_counts.df)
de.diag_vs_early_relapse.celltypes.df <- merge_de_results(de.diag_vs_early_relapse.celltypes)
write.table(de.diag_vs_early_relapse.celltypes.df, 
            paste(work.dir, "de.diag_vs_early_relapse.cell_types.txt", sep = ""), 
            sep = "\t", quote = FALSE, row.names = FALSE)

# diagnostic vs. late relapse
comparators <- sort(unique(hl_all.combined$celltype.earlyrelapse))
# remove the "_earlyRelapse" and RLN cases
comparators <- comparators[ !grepl("_(earlyRelapse|RLN)", comparators)]

# perform the DE analysis
de.diag_vs_late_relapse.celltypes <- get_de_genes(comparators, de.test = "MAST", cell_counts.df)
de.diag_vs_late_relapse.celltypes.df <- merge_de_results(de.diag_vs_late_relapse.celltypes)
write.table(de.diag_vs_late_relapse.celltypes.df, 
            paste(work.dir, "de.diag_vs_late_relapse.cell_types.txt", sep = ""), 
            sep = "\t", quote = FALSE, row.names = FALSE)

## ------ cluster level -------

# define comparators
hl_all.combined$clusters.earlyrelapse <- paste(hl_all.combined$seurat_clusters, hl_all.combined$earlyRelapse, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$clusters.earlyrelapse 
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$clusters.earlyrelapse))

# diagnostic vs. early relapse
comparators <- sort(unique(hl_all.combined$clusters.earlyrelapse))
# remove the "_lateRelapse" and RLN cases
comparators <- comparators[ !grepl("_(lateRelapse|RLN)", comparators)]

# perform the DE analysis
de.diag_vs_early_relapse.clusters <- get_de_genes(comparators, de.test = "MAST", cell_counts.df)
de.diag_vs_early_relapse.clusters.df <- merge_de_results(de.diag_vs_early_relapse.clusters)
write.table(de.diag_vs_early_relapse.clusters.df, 
            paste(work.dir, "de.diag_vs_early_relapse.clusters.txt", sep = ""), 
            sep = "\t", quote = FALSE, row.names = FALSE)

# diagnostic vs. late relapse
comparators <- sort(unique(hl_all.combined$clusters.earlyrelapse))
# remove the "_earlyRelapse" and RLN cases
comparators <- comparators[ !grepl("_(earlyRelapse|RLN)", comparators)]

# perform the DE analysis
de.diag_vs_late_relapse.clusters <- get_de_genes(comparators, de.test = "MAST", cell_counts.df)
de.diag_vs_late_relapse.clusters.df <- merge_de_results(de.diag_vs_late_relapse.clusters)
write.table(de.diag_vs_late_relapse.clusters.df, 
            paste(work.dir, "de.diag_vs_late_relapse.clusters.txt", sep = ""), 
            sep = "\t", quote = FALSE, row.names = FALSE)


## #########################################################
## Cluster-level DE analysis: refractory vs. relapse
## #########################################################

## ------ cell type level -------

# define comparators
Idents(hl_all.combined) <- hl_all.combined$celltype.refractory
comparators <- rev(sort(unique(hl_all.combined$celltype.refractory)))

# remove the "_diagnostic" and RLN cases
comparators <- comparators[ !grepl("_(diagnostic|RLN)", comparators)]
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$celltype.refractory))

# perform the DE analysis
de.refractory_vs_relapse.celltypes <- get_de_genes(comparators, de.test = "MAST", cell_counts.df)
de.refractory_vs_relapse.celltypes.df <- merge_de_results(de.refractory_vs_relapse.celltypes)
write.table(de.refractory_vs_relapse.celltypes.df, 
            paste(work.dir, "de.refractory_vs_relapse.cell_types.txt", sep = ""), 
            sep = "\t", quote = FALSE, row.names = FALSE)

## ------ cluster level -------

# define comparators
Idents(hl_all.combined) <- hl_all.combined$cluster.refractory
comparators <- rev(sort(unique(hl_all.combined$cluster.refractory)))

# remove the "_diagnostic" and RLN cases
comparators <- comparators[ !grepl("_(diagnostic|RLN)", comparators)]
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$cluster.refractory))

# perform the DE analysis
de.refractory_vs_relapse.clusters <- get_de_genes(comparators, de.test = "MAST", cell_counts.df)
de.refractory_vs_relapse.clusters.df <- merge_de_results(de.refractory_vs_relapse.clusters)
write.table(de.refractory_vs_relapse.clusters.df, 
            paste(work.dir, "de.refractory_vs_relapse.clusters.txt", sep = ""), 
            sep = "\t", quote = FALSE, row.names = FALSE)


## #########################################################
## Cluster-level DE analysis: early relapse vs. late relapse
## #########################################################

## ------ cell type level -------

# define comparators
Idents(hl_all.combined) <- hl_all.combined$celltype.earlyrelapse 
comparators <- sort(unique(hl_all.combined$celltype.earlyrelapse))

# remove the "_diagnostic" and RLN cases
comparators <- comparators[ !grepl("_(diagnostic|RLN)", comparators)]
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$celltype.earlyrelapse))

# perform the DE analysis
de.early_vs_late_relapse.celltypes <- get_de_genes(comparators, de.test = "MAST", cell_counts.df)
de.early_vs_late_relapse.celltypes.df <- merge_de_results(de.early_vs_late_relapse.celltypes)
write.table(de.early_vs_late_relapse.celltypes.df, 
            paste(work.dir, "de.early_vs_late_relapse.cell_types.txt", sep = ""),
            sep = "\t", quote = FALSE, row.names = FALSE)

## ------ cluster level -------

# define comparators
Idents(hl_all.combined) <- hl_all.combined$clusters.earlyrelapse 
comparators <- sort(unique(hl_all.combined$clusters.earlyrelapse))

# remove the "_diagnostic" and RLN cases
comparators <- comparators[ !grepl("_(diagnostic|RLN)", comparators)]
# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$clusters.earlyrelapse))

# perform the DE analysis
de.early_vs_late_relapse.clusters <- get_de_genes(comparators, de.test = "MAST", cell_counts.df)
de.early_vs_late_relapse.clusters.df <- merge_de_results(de.early_vs_late_relapse.clusters)
write.table(de.early_vs_late_relapse.clusters.df, 
            paste(work.dir, "de.early_vs_late_relapse.clusters.txt", sep = ""),
            sep = "\t", quote = FALSE, row.names = FALSE)

## #########################################################
## Cluster-level DE analysis: HL (diagnostic + relapse) vs. RLN
## #########################################################

## ------ cell-type level -------

# define comparators (for annotated cell types)
hl_all.combined$celltype.biology <- paste(hl_all.combined$celltype, hl_all.combined$biology, sep = "_")
Idents(hl_all.combined) <- hl_all.combined$celltype.biology
comparators <- sort(unique(hl_all.combined$celltype.biology))

# store number of cells for each comparator
cell_counts.df <- as.data.frame(table(hl_all.combined$celltype.biology))

# perform DE analysis, where odd indices = HL, even indices = RLN
de.hl_vs_rln.celltypes <- get_de_genes(comparators = comparators, de.test = "MAST", cell_counts.df = cell_counts.df)
de.hl_vs_rln.celltypes.df <- merge_de_results(de.hl_vs_rln.celltypes)
write.table(de.hl_vs_rln.celltypes.df, 
            paste(work.dir, "de.hl_vs_rln.cell_types.txt", sep=""),
            sep = "\t", quote = FALSE, row.names = FALSE)

```

## Write out subsetted expression data (useful for generating iTALK input?)

```{r}
library(dplyr)
library(scrattch.io)
library(Seurat)

hl_all.combined <- readRDS("~/Documents/sc_HL_relapse/seurat/output/LATEST/hl_all.combined.annotated.meta_labelled.rds")
# get normalized expression values -- rownames are genes, colnames are cell barcodes
norm.expr <- GetAssayData(object = hl_all.combined[["RNA"]], slot = "data")

# create filtered expression matrix containing only macrophage and t cell clusters
clusters.keep <- c("Macrophage", "Treg proliferation", "Helper T cells", "Treg", "CTL")
# get barcodes for cell type
barcodes.celltype <- as.data.frame(hl_all.combined$celltype)



barcodes.celltype.cluster_filtered <- filter(barcodes.celltype, barcodes.celltype$`hl.both.combined$celltype` %in% clusters.keep)
expr.cluster_filtered <- norm.expr[, colnames(norm.expr) %in% rownames(barcodes.celltype.cluster_filtered)]


# write expr out as matrix
write_dgCMatrix_csv(norm.expr, "~/Documents/sc_HL_relapse/seurat/output/LATEST/iTALK/norm.expr.txt", col1_name = "gene", chunk_size = 100)


# extract expression profiles for macrophage and "T cell" annotated cell barcodes
tcell_cluster_index <- 1
tcell_clusters <- c("Naïve T cells", "CTL", "T regs", "Helper T cells")
macrophage_cluster <- c("Macrophages")
# clusters.keep <- c(macrophage_cluster, tcell_clusters[tcell_cluster_index])
# get normalized expression values -- rownames are genes, colnames are cell barcodes
norm.expr <- GetAssayData(object = hl.both.combined[["RNA"]], slot = "data")
# get barcodes for cell type
barcodes.celltype <- as.data.frame(hl.both.combined$celltype)
# get barcodes for diagnostic / relapse
barcodes.diag_vs_rel <- as.data.frame(hl.both.combined$case.type)
# get barcodes for diagnostic / early relapse
barcodes.diag_vs_early_rel <- as.data.frame(hl.both.combined$earlyRelapse)

# create filtered expression matrix containing only macrophage and t cell clusters
clusters.keep <- c(macrophage_cluster, tcell_clusters)
barcodes.celltype.cluster_filtered <- filter(barcodes.celltype, barcodes.celltype$`hl.both.combined$celltype` %in% clusters.keep)
expr.cluster_filtered <- norm.expr[, colnames(norm.expr) %in% rownames(barcodes.celltype.cluster_filtered)]

## IMPORTANT: before appending columns, will likely need to output as matrix (since current object is dgCMatrix)
write_dgCMatrix_csv(expr.cluster_filtered, "~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/hl_combined.norm_expr.macrophage_t_cell_clusters.txt", col1_name = "gene", chunk_size = 500)
expr.cluster_filtered.df <- t(read.table("~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/hl_combined.norm_expr.macrophage_t_cell_clusters.txt", sep = ",", header = TRUE, stringsAsFactors = FALSE))
cols <- expr.cluster_filtered.df[1, ]
colnames(expr.cluster_filtered.df) <- cols
expr.cluster_filtered.df <- expr.cluster_filtered.df[-1, ]
# rename row names to replace "." with "-" so they can be compared to barcode annotations
rownames(expr.cluster_filtered.df) <- gsub('\\.', '-', rownames(expr.cluster_filtered.df))

# include "cell_type" column
expr.cluster_filtered.df.anno <- merge(expr.cluster_filtered.df, barcodes.celltype, by="row.names")
rownames(expr.cluster_filtered.df.anno) <- expr.cluster_filtered.df.anno$Row.names

# include "compare_group" column indicating diagnostic / relapse OR diagnostic / early relapse
expr.cluster_filtered.df.anno <- merge(expr.cluster_filtered.df.anno, barcodes.diag_vs_rel, by="row.names")
rownames(expr.cluster_filtered.df.anno) <- expr.cluster_filtered.df.anno$Row.names

# write out to ~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/expr/
expr.cluster_filtered.df.anno <- merge(expr.cluster_filtered.df.anno, barcodes.diag_vs_early_rel, by="row.names")
rownames(expr.cluster_filtered.df.anno) <- expr.cluster_filtered.df.anno$Row.names

# clean up the matrix
expr.cluster_filtered.df.anno$Row.names <- NULL
colnames(expr.cluster_filtered.df.anno)[colnames(expr.cluster_filtered.df.anno) == 'Row.names'] <- 'barcode'
colnames(expr.cluster_filtered.df.anno)[colnames(expr.cluster_filtered.df.anno) == 'hl.both.combined$celltype'] <- 'cell_type'
colnames(expr.cluster_filtered.df.anno)[colnames(expr.cluster_filtered.df.anno) == 'hl.both.combined$case.type'] <- 'diag_rel'
colnames(expr.cluster_filtered.df.anno)[colnames(expr.cluster_filtered.df.anno) == 'hl.both.combined$earlyRelapse'] <- 'diag_early_rel'

# write out for input into iTALK
write.table(expr.cluster_filtered.df.anno, "~/Documents/sc_HL_relapse/seurat/output/diagnostic_plus_relapse/hl_combined.norm_expr.macrophage_t_cell_clusters.annotated.txt", sep = "\t")
```

## Visualizing differential expression using Seurat

```{r}
# violin plot split by dataset (primary vs. relapse) per gene showing expression across clusters ("celltype")
# hl.both.combined[["celltype"]] <- Idents(hl.both.combined)
Idents(hl_all.combined) <- hl_all.combined$seurat_clusters

# B and T cells
plots <- VlnPlot(hl_all.combined, features = c("CD3D", "MS4A1"), 
                 split.by = "case.type", group.by = "celltype", pt.size = 0, combine = FALSE)
CombinePlots(plots = plots, ncol = 1)

# CTLA4 and LAG3
plots <- VlnPlot(hl_all.combined, features = c("CTLA4", "LAG3"), 
                 split.by = "case.type", group.by = "celltype", pt.size = 0, combine = FALSE)
CombinePlots(plots = plots, ncol = 1)

# genes of interest: "JUND", PD-L1 --> CD274 and  only found in RNA; 
# CXCL12 (no alias? filtered out early on?)
# CD123 --> IL3RA, CD30 --> TNFRSF8 
genes <- c("PDCD1", "TIGIT", "LAG3", "HAVCR2", "CD274", "CD163", "IL3RA", "GZMA", "GZMB", "GZMK", "PRF1", "IL2RA", "CCR4", "FOXP3", "CXCL13", "CXCR5", "CXCR4", "CXCR3", "CXCL9", "CXCL10", "CXCL12", "IFNG", "IL2", "IL4", "TNFRSF8", "JUND", "HLA-DRB5", "MKI67", "TOX", "CD27", "CD44", "CXCL11", "CD80", "CD86", "CD40", "CCL22", "CD28")
genes.low <- c("CD274", "CD163", "CCR4", "FOXP3", "IL2", "IL4", "TNFRSF8", "CXCL11", "CD80", "CCL22")

genes.diag <- c("CXCL9", "CXCL10", "IFNG")
genes.rel <- c("JUND")
genes.CXC <- c("CXCR3", "CXCR4", "CXCR5", "CXCR9", "CXCL10", "CXCL11")
gene.temp <- c("FOXP3")

# plot genes with low expression
p <- VlnPlot(hl_all.combined, features = gene.temp)

# plot GOI per cluster, split by case type (diagnostic, relapse, RLN)
p.by_case <- VlnPlot(hl_all.combined, features = genes, 
        split.by = "case.type", group.by = "seurat_clusters", 
        pt.size = 0, ncol = 1, log = FALSE, sort = "increasing", 
        assay = "RNA", slot = "data") + xlab("")

p.by_case.low <- VlnPlot(hl_all.combined, features = genes.low,
                         split.by = "case.type", group.by = "seurat_clusters",
                         pt.size = 0.1, ncol = 1, sort = "increasing", log = FALSE, 
                         assay = "RNA", slot = "data") + xlab("")

p.by_early_rel <- VlnPlot(hl_all.combined, features = genes, 
        split.by = "earlyRelapse", group.by = "seurat_clusters", 
        pt.size = 0, ncol = 1, log = FALSE, sort = "increasing", 
        assay = "RNA", slot = "data") + xlab("")

p.by_early_rel.low <- VlnPlot(hl_all.combined, features = genes.low,
                         split.by = "earlyRelapse", group.by = "seurat_clusters",
                         log = FALSE, pt.size = 0.1, ncol = 1, sort = "increasing",
                         assay = "RNA", slot = "data") + xlab("")

# JUND in all clusters (split by diagnostic, relapse, RLN)
p.JUND.clusters <- VlnPlot(hl_all.combined, features = c("JUND"),
                         split.by = "case.type", group.by = "seurat_clusters",
                         log = FALSE, pt.size = 0, ncol = 1, 
                         assay = "RNA", slot = "data") + xlab("")

## Is JUND consistently upregulated in all relapse samples?
p.JUND.samples <- VlnPlot(hl_all.combined, features = c("JUND"),
                         split.by = "case.type", group.by = "orig.ident",
                         log = FALSE, pt.size = 0, ncol = 1, 
                         assay = "RNA", slot = "data") + xlab("")

# can add boxplot on top: VlnPlot(object = seurat_mat_filtered, features = markers_filt[[i]], pt.size = 0)+geom_boxplot(width=0.1,fill="white")

p.temp <- VlnPlot(hl_all.combined, features = c("JUND"),
                         split.by = "earlyRelapse", group.by = "orig.ident",
                         log = FALSE, pt.size = 0, ncol = 1, 
                         assay = "RNA", slot = "data") + xlab("")
```

## Visualize V2 vs V3 differences

```{r}
Idents(hl_all.combined) <- hl_all.combined$chemistry

# read in data
v2_v_v3.de <- read.table("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/DE_analysis/de.V2_v_V3.txt", sep = "\t", header = TRUE)

# extract features that have log2FC < -0.5 and adjusted p-value <= 0.001
v2_v_v3.de.sig <- filter(v2_v_v3.de, v2_v_v3.de$avg_log2FC < -0.5 & v2_v_v3.de$p_val_adj <= 0.001)
v2_v_v3.de.sig_genes <- unique(v2_v_v3.de.sig$gene)
write.table(v2_v_v3.de.sig_genes, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/DE_analysis/v2_v_v3.de.sig_genes.txt", row.names = FALSE, quote = FALSE)

# violin plot for each feature showing each sample coloured by chemistry
pdf("VlnPlot-genes_up_in_V3_vs_V2.pdf", width = 40, height = 60)
VlnPlot(hl_all.combined, features = v2_v_v3.de.sig_genes,
                         split.by = "chemistry", group.by = "orig.ident",
                         log = FALSE, pt.size = 0, ncol = 4, 
                         assay = "RNA", slot = "data") + xlab("")
dev.off()

# Stacy's selected genes (known to be associated with cancer / lymphoma)
sig_genes.GOI <- c("JUND", "ATM", "PTEN", "ARID1B", "KMT2A", "CD81")
pdf("VlnPlot-GOI_up_in_V3_vs_V2.pdf", width = 24, height = 16)
VlnPlot(hl_all.combined, features = sig_genes.GOI,
                         split.by = "chemistry", group.by = "orig.ident",
                         log = FALSE, pt.size = 0, ncol = 2, 
                         assay = "RNA", slot = "data") + xlab("")
dev.off()

# genes of interest for Tomo
sig_genes.GOI <- c("LAG3", "GZMA", "GZMB", "GZMK", "PRF1", "HAVCR2", "IFNG", "PDCD1", "CD80", "CD86", "MKI67", "CXCL13", "CXCR5", "CXCR3", "CD28", "TOX", "CXCL10", "CXCL11", "CCL22", "CD163", "CD274", "IL3RA", "HLA-DRB5", "IGF1R", "CSF1R", "DAB2")
# sig_genes.GOI <- c("LAG3")
pdf("VlnPlot-Tomo_GOI_up_in_V3_vs_V2-no_points.pdf", width = 20, height = 40)
VlnPlot(hl_all.combined, features = sig_genes.GOI,
                         split.by = "chemistry", group.by = "orig.ident",
                         log = FALSE, pt.size = 0, ncol = 2, 
                         assay = "RNA", slot = "data") + xlab("")
dev.off()

# get features that have less than 0.5 FC increase in V3 --> ~800 features
# v2_v_v3.de.low_sig <- filter(v2_v_v3.de, v2_v_v3.de$avg_log2FC > -0.5 & v2_v_v3.de$avg_log2FC < -0.25 &
                           # v2_v_v3.de$p_val_adj <= 0.001)
```

## Violin plots with statistical bars

```{r}
library(ggpubr)
library(dplyr)
library(gridExtra)
library(ggplot2)

#####################################################################################
## Example (manually create violin plot for just GZMB):
#####################################################################################

# Step 1: Get the expression data
data.df <- data.frame(expr = hl_all.combined[["RNA"]]@data["GZMB",], 
                    case = hl_all.combined$case.type, 
                    celltype = hl_all.combined$celltype)

expr.df <- data.df %>% filter(celltype == "CTL", case == "diagnostic" | case == "relapse")

# Step 2: Calculate mean / median
expr.df <- as.data.frame(expr.df %>%
  select(expr, case, celltype) %>%
  group_by(case) %>%
  mutate(mean_expr = mean(expr)))

# Step 3: Plot
p <- ggplot(expr.df, aes(x = case, y = expr)) + 
  geom_violin(aes(fill = mean_expr), trim = TRUE, scale = "width") +
  scale_y_continuous(trans='log2') +
  stat_compare_means(comparisons = list(c("diagnostic", "relapse")), label = "p.format") +
  stat_summary(fun.data = custom_quantile, geom = "boxplot", width=0.2) +
  xlab("") + ylab("")

## Other specific plots of interest:

# diagnostic vs. relapse - CTL (cell type): CD80, CD86
VP_by_gene(gene_list = c("CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/LATEST/violin_with_p_vals/CTL-diag_vs_rel-CD80_CD86", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("CTL"), 
         width=11, height=4, 
         isCluster = FALSE, earlyRelapse = FALSE, biology = FALSE)

# diagnostic vs. early relapse - CTL (cell type): CD80, CD86
VP_by_gene(gene_list = c("CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/LATEST/violin_with_p_vals/CTL-diag_vs_early_rel-CD80_CD86", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("CTL"), 
         width=11, height=4, 
         isCluster = FALSE, earlyRelapse = TRUE, biology = FALSE)

# diagnostic vs. relapse - Helper T (cell type): CD80, CD86
VP_by_gene(gene_list = c("CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/LATEST/violin_with_p_vals/HelperT-diag_vs_rel-CD80_CD86", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("Helper T cells"), 
         width=12, height=3, 
         isCluster = FALSE, earlyRelapse = FALSE, biology = FALSE)

# diagnostic vs. early relapse - Helper T (cell type): CD80, CD86
VP_by_gene(gene_list = c("CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/LATEST/violin_with_p_vals/HelperT-diag_vs_early_rel-CD80_CD86", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("Helper T cells"), 
         width=12, height=3, 
         isCluster = FALSE, earlyRelapse = TRUE, biology = FALSE)

# V2 vs. V3 for significantly up-regulated V3 genes
VP_by_gene_for_chemistry(gene_list = v2_v_v3.de.sig_genes,
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/V2_vs_V3-sig_genes_up_in_V3", 
         comp_grps = list(c("V2", "V3")),
         width = 40, height = 30)

#####################################################################################

# Function: Create gradient breaks using quantiles of data
custom_quantile <- function(x) {
  out <- c(quantile(x, 0.1), quantile(x, 0.25), median(x), quantile(x,0.75), quantile(x, 0.9))
  names(out) <- c("ymin", "lower", "middle", "upper", "ymax")
  return(out)
}

# Function: Generate plots for a given set of genes looking at a specific cell type / cluster and comparison
VP_by_gene <- function(gene_list, 
                       file_name, 
                       comp_grps, 
                       cluster_or_cell_types, 
                       width, 
                       height, 
                       isCluster = FALSE, 
                       earlyRelapse = FALSE, 
                       biology = FALSE){
  
  data.df <- NULL
  expr.df <- NULL

  plot_one_gene <- function(gene){
    
    # Step 1: create expr.df dataset (contains columns: expr, case, celltype/cluster, mean_expr)
    if (isCluster) {
      Idents(hl_all.combined) <- hl_all.combined$seurat_clusters
      # Get the expression data
      if (earlyRelapse) {
        data.df <- data.frame(expr = hl_all.combined[["RNA"]]@data[gene,], 
                              case = hl_all.combined$earlyRelapse,
                              cluster = hl_all.combined$seurat_clusters)
      } else if (biology) {
        data.df <- data.frame(expr = hl_all.combined[["RNA"]]@data[gene,], 
                              case = hl_all.combined$biology,
                              cluster = hl_all.combined$seurat_clusters)
        
      } else {
        data.df <- data.frame(expr = hl_all.combined[["RNA"]]@data[gene,], 
                              case = hl_all.combined$case.type,
                              cluster = hl_all.combined$seurat_clusters)
      }
      expr.df <- data.df %>% filter(cluster %in% cluster_or_cell_types, case %in% comp_grps[[1]])
      
    } else {
      Idents(hl_all.combined) <- hl_all.combined$celltype
      
      # Get the expression data
      if (earlyRelapse) {
        data.df <- data.frame(expr = hl_all.combined[["RNA"]]@data[gene,], 
                              case = hl_all.combined$earlyRelapse, 
                              celltype = hl_all.combined$celltype)
      } else if (biology) {
        data.df <- data.frame(expr = hl_all.combined[["RNA"]]@data[gene,], 
                              case = hl_all.combined$biology, 
                              celltype = hl_all.combined$celltype)
      } else {
        data.df <- data.frame(expr = hl_all.combined[["RNA"]]@data[gene,], 
                              case = hl_all.combined$case.type, 
                              celltype = hl_all.combined$celltype)
      }
      expr.df <- data.df %>% filter(celltype %in% cluster_or_cell_types, case %in% comp_grps[[1]])
    }

    # Step 2: Calculate mean / median
    expr.df <- as.data.frame(expr.df %>%
        select(expr, case) %>%
        group_by(case) %>%
        mutate(mean_expr = mean(expr)))

    # Step 3: Plot (note that we are using the Wilcoxon-rank-sum test (deals with smaller numbers better))
    ggplot(expr.df, aes(x = case, y = expr)) + 
      geom_violin(aes(fill = mean_expr), trim = TRUE, scale = "width") +
      #scale_y_continuous(trans='log2') +
      # Add 10% space on the y-axis above the box plots
      scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
      stat_compare_means(comparisons = comp_grps, label = "p.format") +
      stat_summary(fun.data = custom_quantile, geom = "boxplot", width=0.1) +
      xlab("") + ylab("") + ggtitle(gene) +
      theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) +
      theme(axis.text = element_text(size = 12))
    
  }
  purrr::map(gene_list, plot_one_gene) %>% cowplot::plot_grid(plotlist = .)
  file_name <- paste(file_name, ".png", sep = "")
  ggsave(file_name, width = width, height = height)
}

# violin plots for V2 vs. V3 chemistry

VP_by_gene_for_chemistry <- function(gene_list, 
                       file_name, 
                       comp_grps, 
                       width, 
                       height) {
  
  data.df <- NULL
  expr.df <- NULL

  plot_one_gene <- function(gene){
    
    # Step 1: create expr.df dataset (contains columns: expr, chemistry version, mean_expr)
    Idents(hl_all.combined) <- hl_all.combined$chemistry
    # Get the expression data
    data.df <- data.frame(expr = hl_all.combined[["RNA"]]@data[gene,], 
                              chemistry = hl_all.combined$chemistry)
    expr.df <- data.df %>% filter(chemistry %in% comp_grps[[1]])

    # Step 2: Calculate mean / median
    expr.df <- as.data.frame(expr.df %>%
        select(expr, chemistry) %>%
        group_by(chemistry) %>%
        mutate(mean_expr = mean(expr)))

    # Step 3: Plot (note that we are using the Wilcoxon-rank-sum test (deals with smaller numbers better))
    ggplot(expr.df, aes(x = chemistry, y = expr)) + 
      geom_violin(aes(fill = mean_expr), trim = TRUE, scale = "width") +
      #scale_y_continuous(trans='log2') +
      # Add 10% space on the y-axis above the box plots
      scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
      stat_compare_means(comparisons = comp_grps, label = "p.format") +
      stat_summary(fun.data = custom_quantile, geom = "boxplot", width=0.1) +
      xlab("") + ylab("") + ggtitle(gene) +
      theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) +
      theme(axis.text = element_text(size = 12))
    
  }
  purrr::map(gene_list, plot_one_gene) %>% cowplot::plot_grid(plotlist = .)
  file_name <- paste(file_name, ".png", sep = "")
  ggsave(file_name, width = width, height = height)
}
```

## Latest version of violin plots (coloured by mean value)

```{r}
# # 1. pDC cluster: GZB, CD163, CD123 (-> IL3RA), HLA-DRB5
# VP_by_gene(gene_list = c("GZMB", "CD163", "IL3RA", "HLA-DRB5"),
#          file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/pDC-diag_vs_rel", 
#          comp_grps = list(c("diagnostic", "relapse")), 
#          cluster_or_cell_types = c("Plasmacytoid DCs"), 
#          width=10, height=10)
# 
# # 2. Macrophage cluster: CD163, PD-L1 (-> CD274), CD123 (-> IL3RA), CD80, CD86, CXCL9, CXCL10, CXCL11, CD40, CCL22, TIM3 (-> HAVCR2), HLA-DRB5
# VP_by_gene(gene_list = c("CD163", "CD274", "IL3RA", "CD80", "CD86", "CXCL9", "CXCL10", "CXCL11", "CD40", "CCL22", "HAVCR2", "HLA-DRB5"),
#          file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage-diag_vs_rel", 
#          comp_grps = list(c("diagnostic", "relapse")), 
#          cluster_or_cell_types = c("Macrophage"), 
#          width=20, height=10)

# 2. Macrophage/pDC cluster: CD163, PD-L1 (-> CD274), CD123 (-> IL3RA), CD80, CD86, CXCL9, CXCL10, CXCL11, CD40, CCL22, TIM3 (-> HAVCR2), HLA-DRB5, GZMB
VP_by_gene(gene_list = c("CD163", "CD274", "IL3RA", "CD80", "CD86", "CXCL9", "CXCL10", "CXCL11", "CD40", "CCL22", "HAVCR2", "HLA-DRB5", "GZMB", "IGF1R", "CSF1R", "DAB2"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage_pDC-diag_vs_rel",
         comp_grps = list(c("diagnostic", "relapse")),
         cluster_or_cell_types = c("Macrophage/pDC"),
         width=20, height=10)

# 4. Treg (cell type): LAG3, FOXP3, TIM3 (-> HAVCR2), CXCR3, PD-1 (-> PDCD1), Ki67 (-> MKI67), CD30 (-> TNFRSF8), TOX, CXCR4
VP_by_gene(gene_list = c("LAG3", "FOXP3", "HAVCR2", "CXCR3", "PDCD1", "MKI67", "TNFRSF8", "TOX", "CXCR4", "CD80", "CD86", "TCF7"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/Treg-diag_vs_rel", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("T reg"), 
         width=20, height=10)

# 6. Helper T cell (cell type): LAG3, CXCR5, TIM3 (HAVCR2), CXCR3, PD-1 (PDCD1), Ki67 (MKI67), CD30 (TNFRSF8), CXCL13, TOX, CD28, IL-4 (IL4), CXCR4
VP_by_gene(gene_list = c("LAG3", "CXCR5", "HAVCR2", "CXCR3", "PDCD1", "MKI67", "TNFRSF8", "CXCL13", "TOX", "CD28", "IL4", "CXCR4", "CD80", "CD86", "TCF7"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/HelperT-diag_vs_rel", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("T helper"), 
         width=20, height=10)

# 8. CTL (cell type): GZMA, GZMB, GZMK, PRF1, CXCR3, IFNG, TIM3 (HAVCR2), LAG3, PD1 (PDCD1), CXCR4
VP_by_gene(gene_list = c("GZMA", "GZMB", "GZMK", "PRF1", "CXCR3", "IFNG", "HAVCR2", "LAG3", "PDCD1", "CXCR4", "CD80", "CD86", "TCF7", "TOX"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/CTL-diag_vs_rel", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("CTL"), 
         width=20, height=10)

# 9. NK cells: GZMA, GZMB, GZMK, PRF1, CXCR3, IFNG
VP_by_gene(gene_list = c("GZMA", "GZMB", "GZMK", "PRF1", "CXCR3", "IFNG"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/NK-diag_vs_rel", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("NK cell"), 
         width=20, height=10)

# 3. Treg cluster (cluster 4): LAG3, FOXP3, TIM3 (-> HAVCR2), CXCR3, PD-1 (-> PDCD1), Ki67 (-> MKI67), CD30 (-> TNFRSF8), TOX, CXCR4
VP_by_gene(gene_list = c("LAG3", "FOXP3", "HAVCR2", "CXCR3", "PDCD1", "MKI67", "TNFRSF8", "TOX", "CXCR4", "CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/Treg_cluster4-diag_vs_rel", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("4"), 
         width=20, height=10, 
         isCluster = TRUE)

# 5. Helper T cell (cluster 1): LAG3, CXCR5, TIM3 (HAVCR2), CXCR3, PD-1 (PDCD1), Ki67 (MKI67), CD30 (TNFRSF8), CXCL13, TOX, CD28, IL-4 (IL4), CXCR4
VP_by_gene(gene_list = c("LAG3", "CXCR5", "HAVCR2", "CXCR3", "PDCD1", "MKI67", "TNFRSF8", "CXCL13", "TOX", "CD28", "IL4", "CXCR4", "CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/HelperT_cluster1-diag_vs_rel", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("8"), 
         width=20, height=10, 
         isCluster = TRUE)

# 7. CTL (cluster 8): GZMA, GZMB, GZMK, PRF1, CXCR3, IFNG, TIM3 (HAVCR2), LAG3, PD1 (PDCD1), CXCR4
VP_by_gene(gene_list = c("GZMA", "GZMB", "GZMK", "PRF1", "CXCR3", "IFNG", "HAVCR2", "LAG3", "PDCD1", "CXCR4", "CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/CTL_cluster8-diag_vs_rel", 
         comp_grps = list(c("diagnostic", "relapse")), 
         cluster_or_cell_types = c("7"), 
         width=20, height=10, 
         isCluster = TRUE)


# ############################
# diagnostic VS early relapse
# ############################

# # 1. pDC cluster: GZB, CD163, CD123 (-> IL3RA), HLA-DRB5
# VP_by_gene(gene_list = c("GZMB", "CD163", "IL3RA", "HLA-DRB5"),
#          file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/pDC-diag_vs_early_rel", 
#          comp_grps = list(c("diagnostic", "earlyRelapse")), 
#          cluster_or_cell_types = c("Plasmacytoid DCs"), 
#          width=10, height=10, 
#          earlyRelapse = TRUE)
# 
# # 2. Macrophage cluster: CD163, PD-L1 (-> CD274), CD123 (-> IL3RA), CD80, CD86, CXCL9, CXCL10, CXCL11, CD40, CCL22, TIM3 (-> HAVCR2), HLA-DRB5
# VP_by_gene(gene_list = c("CD163", "CD274", "IL3RA", "CD80", "CD86", "CXCL9", "CXCL10", "CXCL11", "CD40", "CCL22", "HAVCR2", "HLA-DRB5"),
#          file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage-diag_vs_early_rel", 
#          comp_grps = list(c("diagnostic", "earlyRelapse")), 
#          cluster_or_cell_types = c("Macrophage"), 
#          width=20, height=10, 
#          earlyRelapse = TRUE)

VP_by_gene(gene_list = c("CD163", "CD274", "IL3RA", "CD80", "CD86", "CXCL9", "CXCL10", "CXCL11", "CD40", "CCL22", "HAVCR2", "HLA-DRB5", "GZMB", "IGF1R", "CSF1R", "DAB2"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage_pDC-diag_vs_early_rel", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("Macrophage/pDC"), 
         width=20, height=10, 
         earlyRelapse = TRUE)

# 4. Treg (cell type): LAG3, FOXP3, TIM3 (-> HAVCR2), CXCR3, PD-1 (-> PDCD1), Ki67 (-> MKI67), CD30 (-> TNFRSF8), TOX, CXCR4
VP_by_gene(gene_list = c("LAG3", "FOXP3", "HAVCR2", "CXCR3", "PDCD1", "MKI67", "TNFRSF8", "TOX", "CXCR4", "CD80", "CD86", "TCF7"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/Treg-diag_vs_early_rel", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("T reg"), 
         width=20, height=10, 
         earlyRelapse = TRUE)

# 6. Helper T cell (cell type): LAG3, CXCR5, TIM3 (HAVCR2), CXCR3, PD-1 (PDCD1), Ki67 (MKI67), CD30 (TNFRSF8), CXCL13, TOX, CD28, IL-4 (IL4), CXCR4
VP_by_gene(gene_list = c("LAG3", "CXCR5", "HAVCR2", "CXCR3", "PDCD1", "MKI67", "TNFRSF8", "CXCL13", "TOX", "CD28", "IL4", "CXCR4", "CD80", "CD86", "TCF7"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/HelperT-diag_vs_early_rel", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("T helper"), 
         width=20, height=10, 
         earlyRelapse = TRUE)

# 8. CTL (cell type): GZMA, GZMB, GZMK, PRF1, CXCR3, IFNG, TIM3 (HAVCR2), LAG3, PD1 (PDCD1), CXCR4
VP_by_gene(gene_list = c("GZMA", "GZMB", "GZMK", "PRF1", "CXCR3", "IFNG", "HAVCR2", "LAG3", "PDCD1", "CXCR4", "CD80", "CD86", "TCF7", "TOX"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/CTL-diag_vs_early_rel", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("CTL"), 
         width=20, height=10, 
         earlyRelapse = TRUE)

# 9. NK cells: GZMA, GZMB, GZMK, PRF1, CXCR3, IFNG
VP_by_gene(gene_list = c("GZMA", "GZMB", "GZMK", "PRF1", "CXCR3", "IFNG"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/NK-diag_vs_early_rel", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("NK cell"), 
         width=20, height=10, 
         earlyRelapse = TRUE)

# 3. Treg cluster (cluster 4): LAG3, FOXP3, TIM3 (-> HAVCR2), CXCR3, PD-1 (-> PDCD1), Ki67 (-> MKI67), CD30 (-> TNFRSF8), TOX, CXCR4
VP_by_gene(gene_list = c("LAG3", "FOXP3", "HAVCR2", "CXCR3", "PDCD1", "MKI67", "TNFRSF8", "TOX", "CXCR4", "CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/Treg_cluster4-diag_vs_early_rel", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("4"), 
         width=20, height=10, 
         isCluster = TRUE, earlyRelapse = TRUE)

# 5. Helper T cell (cluster 1): LAG3, CXCR5, TIM3 (HAVCR2), CXCR3, PD-1 (PDCD1), Ki67 (MKI67), CD30 (TNFRSF8), CXCL13, TOX, CD28, IL-4 (IL4), CXCR4
VP_by_gene(gene_list = c("LAG3", "CXCR5", "HAVCR2", "CXCR3", "PDCD1", "MKI67", "TNFRSF8", "CXCL13", "TOX", "CD28", "IL4", "CXCR4", "CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/HelperT_cluster1-diag_vs_early_rel", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("8"), 
         width=20, height=10, 
         isCluster = TRUE, earlyRelapse = TRUE)

# 7. CTL (cluster 8): GZMA, GZMB, GZMK, PRF1, CXCR3, IFNG, TIM3 (HAVCR2), LAG3, PD1 (PDCD1), CXCR4
VP_by_gene(gene_list = c("GZMA", "GZMB", "GZMK", "PRF1", "CXCR3", "IFNG", "HAVCR2", "LAG3", "PDCD1", "CXCR4", "CD80", "CD86"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/CTL_cluster8-diag_vs_early_rel", 
         comp_grps = list(c("diagnostic", "earlyRelapse")), 
         cluster_or_cell_types = c("7"), 
         width=20, height=10, 
         isCluster = TRUE, earlyRelapse = TRUE)
  
# ############################
# HL VS RLN
# ############################

# 1. pDC cluster: GZB, CD163, CD123 (-> IL3RA), HLA-DRB5
VP_by_gene(gene_list = c("GZMB", "CD163", "IL3RA", "HLA-DRB5"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage_pDC-HL_vs_RLN", 
         comp_grps = list(c("HL", "RLN")), 
         cluster_or_cell_types = c("Macrophage/pDC"), 
         width=10, height=10, 
         biology = TRUE)

```

## Volcano plots

```{r}
library(EnhancedVolcano)
library(ggplot2)
library(dplyr)
library(plyr)
library(tidyr) # separate

# define working directory:
work.dir <- "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/DE_analysis/"

de.V2_vs_V3.celltypes.df <- read.table(paste(work.dir, "de.V2_vs_V3.cell_types.txt", sep = ""), 
                                         header = TRUE, sep = "\t")
de.diag_vs_rel.celltypes.df <- read.table(paste(work.dir, "de.diag_vs_rel.cell_types.txt", sep = ""), 
                                         header = TRUE, sep = "\t")
de.diag_vs_refractory.celltypes.df <- read.table(paste(work.dir, "de.diag_vs_refractory.cell_types.txt", sep = ""), 
                                                header = TRUE, sep = "\t")
de.diag_vs_early_relapse.celltypes.df <- read.table(paste(work.dir, "de.diag_vs_early_relapse.cell_types.txt", sep = ""), 
                                                   header = TRUE, sep = "\t")
de.diag_vs_late_relapse.celltypes.df <- read.table(paste(work.dir, "de.diag_vs_late_relapse.cell_types.txt", sep = ""),
                                                  header = TRUE, sep = "\t")
de.refractory_vs_relapse.celltypes.df <- read.table(paste(work.dir, "de.refractory_vs_relapse.cell_types.txt", sep = ""),
                                                   header = TRUE, sep = "\t")
de.early_vs_late_relapse.celltypes.df <- read.table(paste(work.dir, "de.early_vs_late_relapse.cell_types.txt", sep = ""),
                                                   header = TRUE, sep = "\t")

results <- list(de.V2_vs_V3.celltypes.df,
                de.diag_vs_rel.celltypes.df,
                de.diag_vs_refractory.celltypes.df,
                de.diag_vs_early_relapse.celltypes.df,
                de.diag_vs_late_relapse.celltypes.df,
                de.refractory_vs_relapse.celltypes.df,
                de.early_vs_late_relapse.celltypes.df)

titles <- c("V2 vs. V3", "diagnostic vs. relapse", "diagnostic vs. refractory", "diagnostic vs. early relapse",
            "diagnostic vs. late relapse", "refractory vs. relapse", "early vs. late relapse")

setwd("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/volcano")

for (i in 1:length(results)) {
  title <- titles[i]
  test.df <- results[[i]]
  
  #split "comparison" column by "_" and take the first split element to be cell_type
  test.df <- separate(data = test.df, col = comparison, into = c("cell_type", "group1"), sep = "\\_")
  test.df$group1 <- NULL
  
  LOGFC_CUTOFF = 0.25
  PVAL_CUTOFF = 0.05
  
  keyvals.colour <- ifelse((abs(test.df$avg_log2FC) > LOGFC_CUTOFF & test.df$p_val_adj < PVAL_CUTOFF), 'firebrick3',
        ifelse((abs(test.df$avg_log2FC) > LOGFC_CUTOFF & test.df$p_val_adj >= PVAL_CUTOFF), 'forestgreen',
               ifelse((abs(test.df$avg_log2FC) < LOGFC_CUTOFF & test.df$p_val_adj < PVAL_CUTOFF), 'dodgerblue3',
          'gray32')))
  keyvals.colour[is.na(keyvals.colour)] <- 'black'
  names(keyvals.colour)[keyvals.colour == 'firebrick3'] <- 'q-val < 0.05; |logFC| > 0.25'
  names(keyvals.colour)[keyvals.colour == 'forestgreen'] <- '|logFC| > 0.25'
  names(keyvals.colour)[keyvals.colour == 'dodgerblue3'] <- 'q-val < 0.05'
  names(keyvals.colour)[keyvals.colour == 'gray32'] <- 'unsignificant'
  
  ## ------ Enhanced volcano plot ------
  
  p.e <- EnhancedVolcano(test.df,
      lab = test.df$gene,
      title = title, subtitle = "",
      x = 'avg_log2FC',
      y = 'p_val_adj', 
      ylim = c(-10, 300), 
      xlim = c(-2, 2.5),
      captionLabSize = 0,
      FCcutoff = LOGFC_CUTOFF,
      pCutoff = PVAL_CUTOFF,
      colCustom = keyvals.colour,
      legendPosition = 'bottom')
  
  pdf(paste("volcano-cell_types-", title, ".pdf", sep = ""), width = 20, height = 11)
  p.e + facet_wrap(~ cell_type, ncol=5) + theme(strip.text.x = element_text(size = 10))
  dev.off()
}

## ------ Custom Enhanced volcano plots ------

## diag vs. rel, diag vs. late relapse - LAG3, PD-1 (PDCD1), PD-L1 (CD274), GZB, CD80, CD86, TIM3 (HAVCR2)
# define different cell-types that will be shaded
GOI <- c('LAG3', 'PDCD1', 'CD274', 'GZMB', 'CD80', 'CD86', 'HAVCR2')
# create custom key-value pairs for different cell-types
# this can be achieved with nested ifelse statements
keyvals.shape <- ifelse(test.df$gene %in% GOI, 17, 1)
keyvals.shape[is.na(keyvals.shape)] <- 1
names(keyvals.shape)[keyvals.shape == 1] <- 'all other genes'
names(keyvals.shape)[keyvals.shape == 17] <- 'GOI'

p.e.0 <- EnhancedVolcano(test.df,
    lab = test.df$gene,
    title = title, subtitle = "",
    selectLab = GOI,
    colAlpha = 0.9,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    pointSize = c(ifelse(test.df$gene %in% GOI, 8, 1)),
    labSize = 5.0,
    captionLabSize = 0,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    FCcutoff = LOGFC_CUTOFF,
    pCutoff = PVAL_CUTOFF,
    shapeCustom = keyvals.shape,
    colCustom = keyvals.colour, 
    legendPosition = 'bottom') 

p.facets.grid <- p.e.0 + facet_wrap(~ cell_type, ncol=5) +
  theme(strip.text.x = element_text(size = 10))
```


## EnhancedVolcano plots for clusters (no mappings)

```{r}
# define working directory:
work.dir <- "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/DE_analysis/"


de.V2_vs_V3.clusters.df <- read.table(paste(work.dir, "de.V2_vs_V3.clusters.txt", sep = ""), 
                                         header = TRUE, sep = "\t")
de.diag_vs_rel.clusters.df <- read.table(paste(work.dir, "de.diag_vs_rel.clusters.txt", sep = ""), 
                                         header = TRUE, sep = "\t")
de.diag_vs_refractory.clusters.df <- read.table(paste(work.dir, "de.diag_vs_refractory.clusters.txt", sep = ""), 
                                                header = TRUE, sep = "\t")
de.diag_vs_early_relapse.clusters.df <- read.table(paste(work.dir, "de.diag_vs_early_relapse.clusters.txt", sep = ""), 
                                                   header = TRUE, sep = "\t")
de.diag_vs_late_relapse.clusters.df <- read.table(paste(work.dir, "de.diag_vs_late_relapse.clusters.txt", sep = ""),
                                                  header = TRUE, sep = "\t")
de.refractory_vs_relapse.clusters.df <- read.table(paste(work.dir, "de.refractory_vs_relapse.clusters.txt", sep = ""),
                                                   header = TRUE, sep = "\t")
de.early_vs_late_relapse.clusters.df <- read.table(paste(work.dir, "de.early_vs_late_relapse.clusters.txt", sep = ""),
                                                   header = TRUE, sep = "\t")

results <- list(de.V2_vs_V3.clusters.df,
              de.diag_vs_rel.clusters.df,
              de.diag_vs_refractory.clusters.df,
              de.diag_vs_early_relapse.clusters.df,
              de.diag_vs_late_relapse.clusters.df,
              de.refractory_vs_relapse.clusters.df,
              de.early_vs_late_relapse.clusters.df)

titles <- c("V2 vs. V3", "diagnostic vs. relapse", "diagnostic vs. refractory", "diagnostic vs. early relapse",
            "diagnostic vs. late relapse", "refractory vs. relapse", "early vs. late relapse")

for (i in 1:length(results)) {
  title <- titles[i]
  test.df <- results[[i]]
  
  #split "comparison" column by "_" and take the first split element to be cell_type
  test.df <- separate(data = test.df, col = comparison, into = c("cluster", "group1"), sep = "\\_")
  test.df$group1 <- NULL
  
  LOGFC_CUTOFF = 0.25
  PVAL_CUTOFF = 0.05
  
  keyvals.colour <- ifelse((abs(test.df$avg_log2FC) > LOGFC_CUTOFF & test.df$p_val_adj < PVAL_CUTOFF), 'firebrick3',
        ifelse((abs(test.df$avg_log2FC) > LOGFC_CUTOFF & test.df$p_val_adj >= PVAL_CUTOFF), 'forestgreen',
               ifelse((abs(test.df$avg_log2FC) < LOGFC_CUTOFF & test.df$p_val_adj < PVAL_CUTOFF), 'dodgerblue3',
          'gray32')))
  keyvals.colour[is.na(keyvals.colour)] <- 'black'
  names(keyvals.colour)[keyvals.colour == 'firebrick3'] <- 'q-val < 0.05; |logFC| > 0.25'
  names(keyvals.colour)[keyvals.colour == 'forestgreen'] <- '|logFC| > 0.25'
  names(keyvals.colour)[keyvals.colour == 'dodgerblue3'] <- 'q-val < 0.05'
  names(keyvals.colour)[keyvals.colour == 'gray32'] <- 'unsignificant'
  
  ## ------ Enhanced volcano plot ------
  
  p.e <- EnhancedVolcano(test.df,
      lab = test.df$gene,
      title = title, subtitle = "",
      x = 'avg_log2FC',
      y = 'p_val_adj', 
      ylim = c(-12, 300), 
      xlim = c(-2, 2.8),
      captionLabSize = 0,
      FCcutoff = LOGFC_CUTOFF,
      pCutoff = PVAL_CUTOFF,
      colCustom = keyvals.colour,
      legendPosition = 'bottom')
  p.grid <- p.e + facet_wrap(~ cluster, ncol=6) + theme(strip.text.x = element_text(size = 10))
  
  filename <- paste("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/volcano/volcano-clusters-", 
                    title, ".pdf", sep = "")
  pdf(filename, width = 20, height = 11)
  p.grid
  dev.off()
  
}

## ------ Custom Enhanced volcano plots ------

## diag vs. rel, diag vs. late relapse - LAG3, PD-1 (PDCD1), PD-L1 (CD274), GZB, CD80, CD86, TIM3 (HAVCR2)
# define different cell-types that will be shaded
GOI <- c('LAG3', 'PDCD1', 'CD274', 'GZMB', 'CD80', 'CD86', 'HAVCR2')
# create custom key-value pairs for different cell-types
# this can be achieved with nested ifelse statements
keyvals.shape <- ifelse(test.df$gene %in% GOI, 17, 1)
keyvals.shape[is.na(keyvals.shape)] <- 1
names(keyvals.shape)[keyvals.shape == 1] <- 'all other genes'
names(keyvals.shape)[keyvals.shape == 17] <- 'GOI'


p.e.0 <- EnhancedVolcano(test.df,
    lab = test.df$gene,
    title = title, subtitle = "",
    selectLab = GOI,
    transcriptLabSize = 0,
    colAlpha = 0.9,
    captionLabSize = 0,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    pointSize = c(ifelse(test.df$gene %in% GOI, 8, 1)),
    labSize = 5.0,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    xlim = c(-1, 1),
    ylim = c(0, 50),
    FCcutoff = LOGFC_CUTOFF,
    pCutoff = PVAL_CUTOFF,
    shapeCustom = keyvals.shape,
    colCustom = keyvals.colour, 
    legendPosition = 'bottom') 

p.facets.grid <- p.e.0 + facet_wrap(~ cluster, ncol=6) +
  theme(strip.text.x = element_text(size = 10))

## diag vs. rel, diag vs. late relapse - CXCL10

p.e.1 <- EnhancedVolcano(test.df,
    lab = test.df$gene,
    title = title, subtitle = "",
    selectLab = c('CXCL10'),
    drawConnectors = TRUE,
    xlim = c(-2,2.5),
    labSize = 6.0,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    FCcutoff = LOGFC_CUTOFF,
    pCutoff = PVAL_CUTOFF,
    colCustom = keyvals.colour,
    legendPosition = 'bottom')

## diag vs. refractory - CXCL13, CXCR4, CXCR6

p.e.2 <- EnhancedVolcano(test.df,
    lab = test.df$gene,
    title = title, subtitle = "",
    selectLab = c('CXCL13', 'CXCR4', 'CXCR6'),
    drawConnectors = TRUE,
    #xlim = c(-2,2.5),
    labSize = 6.0,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    FCcutoff = LOGFC_CUTOFF,
    pCutoff = PVAL_CUTOFF,
    colCustom = keyvals.colour,
    legendPosition = 'bottom')

```

## Cell Cycle Scoring and visualization in UMAP space

```{r}
# retrieve S and G2M genes
s.genes <- Seurat::cc.genes$s.genes
g2m.genes <- Seurat::cc.genes$g2m.genes

# assign cell-cycle scores
hl_all.combined <- CellCycleScoring(hl_all.combined, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

# visualize in umap space
DimPlot(hl_all.combined, 
        reduction = "umap")
        #cols = c("G1"="salmon", "G2M"="green", "S"="royalblue"),
        #group.by = "Phase") # split by cohort (diag vs. relapse)

# ridge plots for marker genes in Proliferation and GCB cell types
RidgePlot(hl_all.combined, 
          features = c("PCNA", "TOP2A", "MCM6", "MKI67"), log = TRUE,
          ncol = 2)

hl_all.combined <- RunPCA(hl_all.combined, features = c(s.genes, g2m.genes))
DimPlot(hl_all.combined)
```

## UpSet Plots

```{r}
library(UpSetR)
library(scrattch.io)
library(reshape2)
library(grid)

work.dir <- "~/Documents/sc_HL_relapse/seurat/output/LATEST/expr/"
print.dir <- "~/Documents/sc_HL_relapse/seurat/figures/LATEST/UpSet/"

norm.expr <- GetAssayData(object = hl_all.combined[["RNA"]], slot = "data")

barcodes.cluster <- as.data.frame(hl_all.combined$seurat_clusters)
barcodes.celltype <- as.data.frame(hl_all.combined$celltype)
barcodes.casetype <- as.data.frame(hl_all.combined$case.type)
barcodes.earlyRelapse <- as.data.frame(hl_all.combined$earlyRelapse)

# Pseudocode:
# Step 1: Extract genes of interest (should be column names)
# Step 2: Extract barcodes of interest (e.g. diagnostic case, macrophage cluster 20) (row names)
# Step 3: map counts / expression values to 0 or 1 (i.e. if > 0 --> 1)

## ###################################################################
## Plot 1: Diagnostic/relapse/early relapse samples with macrophage (cluster 20) expression: CXCL10, CD80, CD163, PD-L1, CCL22
## ###################################################################
#file_prefix <- "diagnostic_cluster20_genelist1.expr"
# file_prefix <- "relapse_cluster20_genelist1.expr"
file_prefix <- "early_relapse_cluster20_genelist1.expr"
file_name <- paste(work.dir, file_prefix, ".txt", sep = "")

# 1. filter for genes
gene_list1 <- c("CXCL10", "CD80", "CD163", "CD274", "CCL22")
expr.genes <- norm.expr[ rownames(norm.expr) %in% gene_list1 , ]

# 2. filter barcodes for cluster / cell type and case type
cluster.to_keep <- c("20")
# casetype.to_keep <- c("diagnostic")
# casetype.to_keep <- c("relapse")
casetype.to_keep <- c("earlyRelapse")
barcodes.cluster_filtered <- filter(barcodes.cluster, barcodes.cluster$`hl_all.combined$seurat_clusters` %in% cluster.to_keep)
#barcodes.casetype_filtered <- filter(barcodes.casetype, barcodes.casetype$`hl_all.combined$case.type` %in% casetype.to_keep)
barcodes.casetype_filtered <- filter(barcodes.earlyRelapse, barcodes.earlyRelapse$`hl_all.combined$earlyRelapse` %in% casetype.to_keep)
barcodes.to_filter <- filter(barcodes.cluster_filtered, rownames(barcodes.cluster_filtered) %in% rownames(barcodes.casetype_filtered))
expr.filtered <- expr.genes[, colnames(expr.genes) %in% rownames(barcodes.to_filter) ]
write_dgCMatrix_csv(expr.filtered, file_name, col1_name = "gene", chunk_size = 1)
expr.df <- read.table(file_name, sep = ",", header = TRUE, row.names = 1)
expr.df.t <- as.data.frame(t(expr.df))
expr.df.t$barcode <- rownames(expr.df.t)

# 3. map expression values > 0 to 1
expr.df.t$CXCL10[expr.df.t$CXCL10 != 0] <- 1
expr.df.t$CD80[expr.df.t$CD80 != 0] <- 1
expr.df.t$CD163[expr.df.t$CD163 != 0] <- 1
expr.df.t$CD274[expr.df.t$CD274 != 0] <- 1
expr.df.t$CCL22[expr.df.t$CCL22 != 0] <- 1

pdf(paste(print.dir, file_prefix, ".pdf", sep = ""), width = 6, height = 4)
upset(expr.df.t, sets = gene_list1, order.by = "freq", mainbar.y.label = "Number of cells") 
dev.off()

## ###################################################################
## Plot 2: Diagnostic/relapse/early relapse samples with Helper T cell (cell type) expression: CD4, PD-1 (PDCD1), LAG3, CXCL13, CD80
## ###################################################################
# file_prefix <- "diagnostic_helperT_genelist2.expr"
# file_prefix <- "relapse_helperT_genelist2.expr"
file_prefix <- "early_relapse_helperT_genelist2.expr"
file_name <- paste(work.dir, file_prefix, ".txt", sep = "")

# 1. filter for genes
gene_list2 <- c("CD4", "PDCD1", "LAG3", "CXCL13", "CD80")
expr.genes <- norm.expr[ rownames(norm.expr) %in% gene_list2 , ]

# 2. filter barcodes for cluster / cell type and case type
celltype.to_keep <- c("Helper T cells")
# casetype.to_keep <- c("diagnostic")
# casetype.to_keep <- c("relapse")
casetype.to_keep <- c("earlyRelapse")
barcodes.celltype_filtered <- filter(barcodes.celltype, barcodes.celltype$`hl_all.combined$celltype` %in% celltype.to_keep)
# barcodes.casetype_filtered <- filter(barcodes.casetype, barcodes.casetype$`hl_all.combined$case.type` %in% casetype.to_keep)
barcodes.casetype_filtered <- filter(barcodes.earlyRelapse, barcodes.earlyRelapse$`hl_all.combined$earlyRelapse` %in% casetype.to_keep)
barcodes.to_filter <- filter(barcodes.celltype_filtered, rownames(barcodes.celltype_filtered) %in% rownames(barcodes.casetype_filtered))
expr.filtered <- expr.genes[, colnames(expr.genes) %in% rownames(barcodes.to_filter) ]
write_dgCMatrix_csv(expr.filtered, file_name, col1_name = "gene", chunk_size = 1)
expr.df <- read.table(file_name, sep = ",", header = TRUE, row.names = 1)
expr.df.t <- as.data.frame(t(expr.df))
expr.df.t$barcode <- rownames(expr.df.t)

# 3. map expression values > 0 to 1
expr.df.t$CD4[expr.df.t$CD4 != 0] <- 1
expr.df.t$PDCD1[expr.df.t$PDCD1 != 0] <- 1
expr.df.t$LAG3[expr.df.t$LAG3 != 0] <- 1
expr.df.t$CXCL13[expr.df.t$CXCL13 != 0] <- 1
expr.df.t$CD80[expr.df.t$CD80 != 0] <- 1

pdf(paste(print.dir, file_prefix, ".pdf", sep = ""), width = 6, height = 4)
upset(expr.df.t, sets = gene_list2, order.by = "freq", mainbar.y.label = "Number of cells", text.scale = 0.8) 
dev.off()

## ###################################################################
## Plot 3: Diagnostic/relapse/early relapse samples with Treg (cell type) expression: CD4, PD-1 (PDCD1), LAG3, FOXP3, CD80
## ###################################################################
# file_prefix <- "diagnostic_Treg_genelist3.expr"
# file_prefix <- "relapse_Treg_genelist3.expr"
file_prefix <- "early_relapse_Treg_genelist3.expr"
file_name <- paste(work.dir, file_prefix, ".txt", sep = "")

# 1. filter for genes
gene_list3 <- c("CD4", "PDCD1", "LAG3", "FOXP3", "CD80")
expr.genes <- norm.expr[ rownames(norm.expr) %in% gene_list3 , ]

# 2. filter barcodes for cluster / cell type and case type
celltype.to_keep <- c("Treg")
# casetype.to_keep <- c("diagnostic")
# casetype.to_keep <- c("relapse")
casetype.to_keep <- c("earlyRelapse")
barcodes.celltype_filtered <- filter(barcodes.celltype, barcodes.celltype$`hl_all.combined$celltype` %in% celltype.to_keep)
# barcodes.casetype_filtered <- filter(barcodes.casetype, barcodes.casetype$`hl_all.combined$case.type` %in% casetype.to_keep)
barcodes.casetype_filtered <- filter(barcodes.earlyRelapse, barcodes.earlyRelapse$`hl_all.combined$earlyRelapse` %in% casetype.to_keep)
barcodes.to_filter <- filter(barcodes.celltype_filtered, rownames(barcodes.celltype_filtered) %in% rownames(barcodes.casetype_filtered))
expr.filtered <- expr.genes[, colnames(expr.genes) %in% rownames(barcodes.to_filter) ]
write_dgCMatrix_csv(expr.filtered, file_name, col1_name = "gene", chunk_size = 1)
expr.df <- read.table(file_name, sep = ",", header = TRUE, row.names = 1)
expr.df.t <- as.data.frame(t(expr.df))
expr.df.t$barcode <- rownames(expr.df.t)

# 3. map expression values > 0 to 1
expr.df.t$CD4[expr.df.t$CD4 != 0] <- 1
expr.df.t$PDCD1[expr.df.t$PDCD1 != 0] <- 1
expr.df.t$LAG3[expr.df.t$LAG3 != 0] <- 1
expr.df.t$FOXP3[expr.df.t$FOXP3 != 0] <- 1
expr.df.t$CD80[expr.df.t$CD80 != 0] <- 1

pdf(paste(print.dir, file_prefix, ".pdf", sep = ""), width = 6, height = 4)
upset(expr.df.t, sets = gene_list3, order.by = "freq", mainbar.y.label = "Number of cells", text.scale = 0.8) 
dev.off()

```

## Special request from Richard Corbett - what are the cell types for each barcode?

```{r}
library(reshape2)

# Retrieve cluster assignments (identities) for cell barcodes in cHL-LSARP-09, 10, 11, 12, 19, 20, 25, and 26
cluster.mappings <- read.table("~/Documents/sc_HL_relapse/data/relapse/hl_relapse-cluster_annotations.txt", sep = "\t", header = TRUE)
gsc.data <- read.table("~/Documents/sc_HL_relapse/data/relapse/library_info-relapse-cohort.txt", sep = "\t", header = TRUE, fill = TRUE)

# merge dataframes to combine barcode, cluster id, and cell type annotation
colnames(clusters.df) <- c("barcode_sample", "cluster")
clusters.annotated <- merge(clusters.df, cluster.mappings, by = "cluster")

# extract sample id so we can more easily filter for samples of interest
clusters.annotated <- cbind(clusters.annotated, colsplit(clusters.annotated$barcode_sample, pattern = "_", names = c("barcode", "sample_id")))

# save for future use
write.table(clusters.annotated, "~/Documents/sc_HL_relapse/seurat/output/relapse-all_25_merged/hl_relapse-merged.sct-clusters.annotated.txt", sep = "\t", row.names = FALSE, quote = FALSE)

# extract the essential columns
keeps <- c("sample_id", "barcode", "cluster", "cell_type")
clusters.annotated <- clusters.annotated[ , keeps ]

# now extract samples of interest
samples.keep <- c("cHL-LSARP-09", "cHL-LSARP-10", "cHL-LSARP-11", "cHL-LSARP-12",
                  "cHL-LSARP-19", "cHL-LSARP-20", "cHL-LSARP-25", "cHL-LSARP-26")
clusters.annotated.RC <- filter(clusters.annotated, clusters.annotated$sample_id %in% samples.keep)

# sort by sample id then cluster
clusters.annotated.RC <- clusters.annotated.RC[order(clusters.annotated.RC$sample_id, clusters.annotated.RC$cluster), ]

# add library information
gsc.library <- gsc.data[ c("Sample", "Library", "GSCPool") ]
colnames(gsc.library) <- c("sample_id", "gsc_library", "gsc_pool")
clusters.annotated.RC <- merge(clusters.annotated.RC, gsc.library, by = "sample_id")
keeps <- c("sample_id", "gsc_library", "gsc_pool", "barcode", "cluster", "cell_type")
clusters.annotated.RC <- clusters.annotated.RC[ , keeps]

# write to output
write.table(clusters.annotated.RC, "~/Documents/sc_HL_relapse/for_Richard/lsarp_hl_relapse-annotations-RC_samples.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

