---
title: "LSARP_analysis-CD4_CTL_clustering"
author: "Stacy Hung"
date: "20/01/2021"
output: html_document
---

## Load libraries

```{r}
library(Seurat)
library(dplyr)
library(pheatmap)
```

## Step 1: Filter for B cells based on expression / non-expression of marker combinations

```{r}
hl_all.combined <- readRDS("~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.integrated.annotated.rds")

# Step 1: Filter cells for B cells based on:
#   Expression of CD20 (MS4A1)
#   No expression of CD3 (CD3D) CD4, CD8 (CD8A), CD68, CD56 (NCAM1)
hl.B_markers.cells <- which(hl_all.combined@assays$RNA@data["MS4A1",] > 0 & 
                               hl_all.combined@assays$RNA@data["CD3D",] == 0 &
                               hl_all.combined@assays$RNA@data["CD4",] == 0 &
                               hl_all.combined@assays$RNA@data["CD8A",] == 0 &
                               hl_all.combined@assays$RNA@data["NCAM1",] == 0 &
                               hl_all.combined@assays$RNA@data["CD68",] == 0
                            ) # ? (?% of all cells)
hl.filtered.B_markers <- subset(hl_all.combined, cells = hl.B_markers.cells)
```

## Step 2: Subcluster and plot heatmaps of clusters

```{r}
DefaultAssay(hl.filtered.B_markers) <- "integrated"
hl.filtered.B_markers <- FindNeighbors(hl.filtered.B_markers, reduction = "pca", dims = 1:30)
hl.filtered.B_markers.subclustered <- FindClusters(hl.filtered.B_markers, resolution = 0.4)

Idents(hl.filtered.B_markers.subclustered) <- hl.filtered.B_markers.subclustered$seurat_clusters

# plot umap
pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/BReg/umap-Breg-subclusters.pdf", width = 6, height = 5)
DimPlot(hl.filtered.B_markers.subclustered, reduction = "umap", label = TRUE) 
dev.off()

# plot average expression heatmap for component markers
avg_exp.heatmaps <- avgComponentExpressionHeatmap(data = hl.filtered.B_markers.subclustered)

# plot average expression heatmap for *B-cell-specific* markers
avg_exp.heatmaps.B_cell <- avg_B_ComponentExpressionHeatmap(data = hl.filtered.B_markers.subclustered)

# plot individual expression heatmaps for each cluster
printSCHeatmapsPerCluster(data = hl.filtered.B_markers.subclustered, 
                          plot_folder = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/BReg/")

printSCHeatmapsPerCluster_Bmarkers(data = hl.filtered.B_markers.subclustered, 
                          plot_folder = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/BReg/")
```

## Annotate clusters (Tomo)

```{r}
# attach manual annotations
hl.filtered.B_markers.subclustered <- RenameIdents(hl.filtered.B_markers.subclustered,
                                `0` = "pDC",
                                `1` = "macrophage",
                                `2` = "macrophage",
                                `3` = "pDC",
                                `4` = "macrophage")
hl.filtered.B_markers.subclustered$celltype2 <- Idents(hl.filtered.B_markers.subclustered)
pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/CD4/umap-CD4-subclusters-annotated.pdf", width = 6, height = 5)
DimPlot(hl.filtered.B_markers.subclustered, reduction = "umap", label = TRUE) 
dev.off()

saveRDS(hl.filtered.B_markers.subclustered, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.filtered.B_markers.subclustered.annotated.rds")
```

## Function: Create heatmap of average expression of clusters across *B cell* component markers

Parameters:
  - data: Seurat object with $seurat_clusters

```{r}
avg_B_ComponentExpressionHeatmap <- function(data) {
  # get component markers
  component.mappings <- read.table("~/Documents/sc_HL_relapse/data/B_component_genes.txt", 
                                   sep = "\t", header = TRUE, fill = TRUE)
  # first, get list of highly-variable genes
  hvg <- data@assays$integrated@var.features
  # only look at markers that are in the HVG list
  markers.to.plot <- subset(hvg, hvg %in% component.mappings$Gene)
  # get average expression of each gene
  cluster.averages <- AverageExpression(data, features = markers.to.plot, return.seurat = TRUE)
  #data.avg <- GetAssayData(object = cluster.averages, slot = "scale.data")
  data.avg <- GetAssayData(object = cluster.averages, slot = "data", assay = "SCT")
  # now match component gene list to those being plotted
  component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
  rownames(component.mappings) <- component.mappings$Gene
  # order by component then gene name so heatmap legend makes sense
  components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
  component.mappings$Gene <- NULL
  component.mappings$Alt_name <- NULL
  # reorder expression matrix to match component mappings
  data.avg.ordered <- data.avg[match(rownames(components.ordered), rownames(data.avg)), ]

  heatmaps <- vector('list', 2)
  
  # heatmap of average expression per cluster across all component genes
  heatmaps[[1]] <- pheatmap(data.avg.ordered, 
         annotation_row = component.mappings,
         cluster_rows = FALSE, 
         annotation_names_row = FALSE, 
         gaps_row = c(2,3,9,10,12,13), 
         angle_col = 45)
  # rotated version of heatmap
  heatmaps[[2]] <- pheatmap(t(data.avg.ordered), 
         annotation_col = component.mappings,
         cluster_cols = FALSE, 
         annotation_names_col = FALSE, 
         gaps_col = c(2,3,9,10,12,13), 
         angle_col = 45)
  
  return(heatmaps)
}

```


## Function: Create heatmap of average expression of clusters across component markers

Parameters:
  - data: Seurat object with $seurat_clusters

```{r}
avgComponentExpressionHeatmap <- function(data) {
  # get component markers
  component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", 
                                   sep = "\t", header = TRUE, fill = TRUE)
  # first, get list of highly-variable genes
  hvg <- data@assays$integrated@var.features
  # only look at markers that are in the HVG list
  markers.to.plot <- subset(hvg, hvg %in% component.mappings$Gene)
  # get average expression of each gene
  cluster.averages <- AverageExpression(data, features = markers.to.plot, return.seurat = TRUE)
  #data.avg <- GetAssayData(object = cluster.averages, slot = "scale.data")
  data.avg <- GetAssayData(object = cluster.averages, slot = "data", assay = "SCT")
  # now match component gene list to those being plotted
  component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
  rownames(component.mappings) <- component.mappings$Gene
  # order by component then gene name so heatmap legend makes sense
  components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
  component.mappings$Gene <- NULL
  component.mappings$Alt_name <- NULL
  # reorder expression matrix to match component mappings
  data.avg.ordered <- data.avg[match(rownames(components.ordered), rownames(data.avg)), ]

  heatmaps <- vector('list', 2)
  
  # heatmap of average expression per cluster across all component genes
  heatmaps[[1]] <- pheatmap(data.avg.ordered, 
         annotation_row = component.mappings,
         cluster_rows = FALSE, 
         annotation_names_row = FALSE, 
         angle_col = 45, 
         gaps_row = c(14,20,22,24,27,32,34,38,39,40,41,44,51,53))
  # rotated version of heatmap
  heatmaps[[2]] <- pheatmap(t(data.avg.ordered), 
         annotation_col = component.mappings,
         cluster_cols = FALSE, 
         annotation_names_col = FALSE, 
         gaps_col = c(14,20,22,24,27,32,34,38,39,40,41,44,51,53), 
         angle_col = 45)
  
  return(heatmaps)
}

```

## Function: Produce single-cell level heatmaps of expression for EACH cluster using *B cell* markers

Parameters:
  - data: Seurat object with $seurat_clusters
  - plot_folder: absolute path of folder to print heatmaps to

```{r}
printSCHeatmapsPerCluster_Bmarkers <- function(data, plot_folder) {
  # get component markers
  component.mappings <- read.table("~/Documents/sc_HL_relapse/data/B_component_genes.txt", 
                                   sep = "\t", header = TRUE, fill = TRUE)
  # first, get list of highly-variable genes
  hvg <- data@assays$integrated@var.features
  # only look at markers that are in the HVG list
  markers.to.plot <- subset(hvg, hvg %in% component.mappings$Gene)
  # get total number of clusters
  NUM_CLUSTERS = length(unique(data$seurat_clusters))
  # get expression matrix (features by barcodes)
  data.expr <- data@assays$SCT@data
  # retrieve expression for each cell across the component genes of interest
  data.expr.markers <- data.expr[ rownames(data.expr) %in% markers.to.plot , ]
  # match component gene list to those that can be plotted
  component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
  # order by component then gene name so that the heatmap legend makes sense
  rownames(component.mappings) <- component.mappings$Gene
  components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
  components.ordered$Gene <- NULL
  components.ordered$Alt_name <- NULL
  # reorder expression matrix to match component mappings
  data.expr.markers.ordered <- data.expr.markers[ row.names(components.ordered), ]
  # to create separate expr matrices for each cluster, need mappings for cell --> cluster
  data.clusters <- data$seurat_clusters
  # convert to char
  clusters.char <- levels(data.clusters)[data.clusters] # ordered vector of cluster ids (0 through 26)
  barcodes.cluster <- attributes(data.clusters)$names # ordered vector of barcodes (76,607 unique entries)
  clusters.df <- data.frame(barcodes.cluster, clusters.char)
 
  expr.markers <- vector('list', NUM_CLUSTERS)
  heatmaps <- vector('list', NUM_CLUSTERS)
  hl.seurat <- vector('list', NUM_CLUSTERS)
  
  for (c in 1:NUM_CLUSTERS) {
    hl.seurat[[c]] <- subset(data, idents = c-1) # filter by cluster
    hl.seurat[[c]] <- subset(hl.seurat[[c]], features = markers.to.plot) # filter by genes
    
    # retrieve expression for each cell across the component genes of interest
    expr.temp <- hl.seurat[[c]]@assays$SCT@data
    expr.markers[[c]] <- expr.temp[ rownames(expr.temp) %in% markers.to.plot , ]
    
    # reorder expression matrix to match component mappings
    expr.markers[[c]] <- expr.markers[[c]][ row.names(components.ordered), ]
  
    filename <- paste(plot_folder, "heatmap-B_markers-cluster", 
                      c-1, ".pdf", sep = "")
    pdf(filename, width = 8, height = 4)
    heatmaps[[c]] <- pheatmap(expr.markers[[c]], 
           annotation_row = components.ordered, 
           cluster_rows = FALSE, 
           annotation_names_row = FALSE, 
           angle_col = 45, 
           gaps_row = c(2,3,9,10,12,13), 
           border_color = NA, show_colnames = FALSE, scale = "none")
    dev.off()
  }
 
}
```

## Function: Produce single-cell level heatmaps of expression for EACH cluster

Parameters:
  - data: Seurat object with $seurat_clusters
  - plot_folder: absolute path of folder to print heatmaps to

```{r}
printSCHeatmapsPerCluster <- function(data, plot_folder) {
  # get component markers
  component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", 
                                   sep = "\t", header = TRUE, fill = TRUE)
  # first, get list of highly-variable genes
  hvg <- data@assays$integrated@var.features
  # only look at markers that are in the HVG list
  markers.to.plot <- subset(hvg, hvg %in% component.mappings$Gene)
  # get total number of clusters
  NUM_CLUSTERS = length(unique(data$seurat_clusters))
  # get expression matrix (features by barcodes)
  data.expr <- data@assays$SCT@data
  # retrieve expression for each cell across the component genes of interest
  data.expr.markers <- data.expr[ rownames(data.expr) %in% markers.to.plot , ]
  # match component gene list to those that can be plotted
  component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
  # order by component then gene name so that the heatmap legend makes sense
  rownames(component.mappings) <- component.mappings$Gene
  components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
  components.ordered$Gene <- NULL
  components.ordered$Alt_name <- NULL
  # reorder expression matrix to match component mappings
  data.expr.markers.ordered <- data.expr.markers[ row.names(components.ordered), ]
  # to create separate expr matrices for each cluster, need mappings for cell --> cluster
  data.clusters <- data$seurat_clusters
  # convert to char
  clusters.char <- levels(data.clusters)[data.clusters] # ordered vector of cluster ids (0 through 26)
  barcodes.cluster <- attributes(data.clusters)$names # ordered vector of barcodes (76,607 unique entries)
  clusters.df <- data.frame(barcodes.cluster, clusters.char)
 
  expr.markers <- vector('list', NUM_CLUSTERS)
  heatmaps <- vector('list', NUM_CLUSTERS)
  hl.seurat <- vector('list', NUM_CLUSTERS)
  
  for (c in 1:NUM_CLUSTERS) {
    hl.seurat[[c]] <- subset(data, idents = c-1) # filter by cluster
    hl.seurat[[c]] <- subset(hl.seurat[[c]], features = markers.to.plot) # filter by genes
    
    # retrieve expression for each cell across the component genes of interest
    expr.temp <- hl.seurat[[c]]@assays$SCT@data
    expr.markers[[c]] <- expr.temp[ rownames(expr.temp) %in% markers.to.plot , ]
    
    # reorder expression matrix to match component mappings
    expr.markers[[c]] <- expr.markers[[c]][ row.names(components.ordered), ]
  
    filename <- paste(plot_folder, "heatmap-cluster", 
                      c-1, ".pdf", sep = "")
    pdf(filename, width = 8, height = 8)
    heatmaps[[c]] <- pheatmap(expr.markers[[c]], 
           annotation_row = components.ordered, 
           cluster_rows = FALSE, 
           annotation_names_row = FALSE, 
           angle_col = 45, 
           gaps_row = c(14,20,22,24,27,32,34,38,39,40,41,44,51,53), 
           border_color = NA, show_colnames = FALSE, scale = "none")
    dev.off()
  }
 
}
```

