---
title: "LSARP_analysis-cluster_annotation"
author: "Stacy Hung"
date: "25/11/2020"
output: html_document
---

## Subcluster macrophage / pDC cluster

```{r}
library(Seurat)

DefaultAssay(hl.macrophage_pDC) <- "integrated"
hl.macrophage_pDC <- FindNeighbors(hl.macrophage_pDC, reduction = "pca", dims = 1:30)
hl.macrophage_pDC.subclustered <- FindClusters(hl.macrophage_pDC, resolution = 0.4)

Idents(hl.macrophage_pDC.subclustered) <- hl.macrophage_pDC.subclustered$seurat_clusters

# plot umap
pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/umap-macrophage_pDC-subclusters.pdf", width = 6, height = 5)
DimPlot(hl.macrophage_pDC.subclustered, reduction = "umap", label = TRUE) 
dev.off()

# plot average expression heatmap for component markers
avgComponentExpressionHeatmap(data = hl.macrophage_pDC.subclustered)
# plot individual expression heatmaps for each cluster
printSCHeatmapsPerCluster(data = hl.macrophage_pDC.subclustered, 
                          plot_folder = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/subclustering/")
# attach manual annotations
hl.macrophage_pDC.subclustered <- RenameIdents(hl.macrophage_pDC.subclustered,
                                `0` = "pDC",
                                `1` = "macrophage",
                                `2` = "macrophage",
                                `3` = "pDC",
                                `4` = "macrophage")
hl.macrophage_pDC.subclustered$celltype2 <- Idents(hl.macrophage_pDC.subclustered)
pdf("~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/umap-macrophage_pDC-subclusters-annotated.pdf", width = 6, height = 5)
DimPlot(hl.macrophage_pDC.subclustered, reduction = "umap", label = TRUE) 
dev.off()

saveRDS(hl.macrophage_pDC.subclustered, "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/hl.macrophage_pDC.subclustered.annotated.rds")
```

## DE analysis

```{r}
library(MAST)

work_dir <- "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/DE_analysis/macrophage_pDC_cluster/"

# diagnostic vs. relapse
hl.macrophage_pDC.subclustered$comparator <- paste(hl.macrophage_pDC.subclustered$celltype2,
                                                       hl.macrophage_pDC.subclustered$case.type, sep = "_")
writeDEresults(data = hl.macrophage_pDC.subclustered, 
               grep_string = "_RLN", 
               file_name = "de.diag_vs_rel.cell_types.txt")

# diagnostic vs. refractory
hl.macrophage_pDC.subclustered$comparator <- paste(hl.macrophage_pDC.subclustered$celltype2,
                                                   hl.macrophage_pDC.subclustered$refractory, sep = "_")
writeDEresults(data = hl.macrophage_pDC.subclustered, 
               grep_string = "_(relapseNotRefractory|RLN)", 
               file_name = "de.diag_vs_refractory.cell_types.txt")

# refractory vs. relapse
writeDEresults(data = hl.macrophage_pDC.subclustered, 
               grep_string = "_(diagnostic|RLN)", 
               file_name = "de.refractory_vs_relapse.cell_types.txt")

# diagnostic vs. early relapse
hl.macrophage_pDC.subclustered$comparator <- paste(hl.macrophage_pDC.subclustered$celltype2,
                                                   hl.macrophage_pDC.subclustered$earlyRelapse, sep = "_")
writeDEresults(data = hl.macrophage_pDC.subclustered, 
               grep_string = "_(lateRelapse|RLN)", 
               file_name = "de.diag_vs_early_relapse.cell_types.txt")

# diagnostic vs. late relapse
writeDEresults(data = hl.macrophage_pDC.subclustered, 
               grep_string = "_(earlyRelapse|RLN)", 
               file_name = "de.diag_vs_late_relapse.cell_types.txt")

# early vs. late relapse
writeDEresults(data = hl.macrophage_pDC.subclustered, 
               grep_string = "_(diagnostic|RLN)", 
               file_name = "de.early_vs_late_relapse.cell_types.txt")

```

## Volcano plots

```{r}
library(EnhancedVolcano)
library(ggplot2)
library(dplyr)
library(plyr)
library(tidyr) # separate

# define working directory:
work.dir <- "~/Documents/sc_HL_relapse/seurat/output/V4_LATEST/DE_analysis/macrophage_pDC_cluster/"

de.diag_vs_rel.celltypes.df <- read.table(paste(work.dir, "de.diag_vs_rel.cell_types.txt", sep = ""), 
                                         header = TRUE, sep = "\t")
de.diag_vs_refractory.celltypes.df <- read.table(paste(work.dir, "de.diag_vs_refractory.cell_types.txt", sep = ""), 
                                                header = TRUE, sep = "\t")
de.diag_vs_early_relapse.celltypes.df <- read.table(paste(work.dir, "de.diag_vs_early_relapse.cell_types.txt", sep = ""), 
                                                   header = TRUE, sep = "\t")
de.diag_vs_late_relapse.celltypes.df <- read.table(paste(work.dir, "de.diag_vs_late_relapse.cell_types.txt", sep = ""),
                                                  header = TRUE, sep = "\t")
de.refractory_vs_relapse.celltypes.df <- read.table(paste(work.dir, "de.refractory_vs_relapse.cell_types.txt", sep = ""),
                                                   header = TRUE, sep = "\t")
de.early_vs_late_relapse.celltypes.df <- read.table(paste(work.dir, "de.early_vs_late_relapse.cell_types.txt", sep = ""),
                                                   header = TRUE, sep = "\t")

results <- list(de.diag_vs_rel.celltypes.df,
                de.diag_vs_refractory.celltypes.df,
                de.diag_vs_early_relapse.celltypes.df,
                de.diag_vs_late_relapse.celltypes.df,
                de.refractory_vs_relapse.celltypes.df,
                de.early_vs_late_relapse.celltypes.df)

titles <- c("diagnostic vs. relapse", "diagnostic vs. refractory", "diagnostic vs. early relapse",
            "diagnostic vs. late relapse", "refractory vs. relapse", "early vs. late relapse")

dir <- "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/volcano/macrophage_pDC_subcluster/"

for (i in 1:length(results)) {
  title <- titles[i]
  test.df <- results[[i]]
  
  #split "comparison" column by "_" and take the first split element to be cell_type
  test.df <- separate(data = test.df, col = comparison, into = c("cell_type", "group1"), sep = "\\_")
  test.df$group1 <- NULL
  
  LOGFC_CUTOFF = 0.25
  PVAL_CUTOFF = 0.05
  
  keyvals.colour <- ifelse((abs(test.df$avg_log2FC) > LOGFC_CUTOFF & test.df$p_val_adj < PVAL_CUTOFF), 'firebrick3',
        ifelse((abs(test.df$avg_log2FC) > LOGFC_CUTOFF & test.df$p_val_adj >= PVAL_CUTOFF), 'forestgreen',
               ifelse((abs(test.df$avg_log2FC) < LOGFC_CUTOFF & test.df$p_val_adj < PVAL_CUTOFF), 'dodgerblue3',
          'gray32')))
  keyvals.colour[is.na(keyvals.colour)] <- 'black'
  names(keyvals.colour)[keyvals.colour == 'firebrick3'] <- 'q-val < 0.05; |logFC| > 0.25'
  names(keyvals.colour)[keyvals.colour == 'forestgreen'] <- '|logFC| > 0.25'
  names(keyvals.colour)[keyvals.colour == 'dodgerblue3'] <- 'q-val < 0.05'
  names(keyvals.colour)[keyvals.colour == 'gray32'] <- 'unsignificant'
  
  ## ------ Enhanced volcano plot ------
  
  p.e <- EnhancedVolcano(test.df,
      lab = test.df$gene,
      title = title, subtitle = "",
      x = 'avg_log2FC',
      y = 'p_val_adj', 
      ylim = c(-1, 15), 
      xlim = c(-3, 3),
      captionLabSize = 0,
      FCcutoff = LOGFC_CUTOFF,
      pCutoff = PVAL_CUTOFF,
      colCustom = keyvals.colour,
      legendPosition = 'bottom')
  
  pdf(paste(dir, "volcano-cell_types-", title, ".pdf", sep = ""), width = 6, height = 6)
  p.e + facet_wrap(~ cell_type, ncol=5) + theme(strip.text.x = element_text(size = 10))
  dev.off()
}
```

## Violin plots

```{r}
# diagnostic vs. relapse
Idents(hl.macrophage_pDC.subclustered) <- hl.macrophage_pDC.subclustered$celltype2

# macrophage
VP_by_gene(data = hl.macrophage_pDC.subclustered,
           gene_list = c("CD163", "CD274", "IL3RA", "CD80", "CD86", "CXCL9", "CXCL10", "CXCL11", "CD40", "CCL22", "HAVCR2", "HLA-DRB5", "GZMB", "IGF1R", "CSF1R", "DAB2"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage_pDC_subcluster/macrophage-diag_vs_rel",
         comp_grps = list(c("diagnostic", "relapse")),
         cluster_or_cell_types = c("macrophage"),
         width=20, height=10)

# pDC
VP_by_gene(data = hl.macrophage_pDC.subclustered,
           gene_list = c("CD163", "CD274", "IL3RA", "CD80", "CD86", "CXCL9", "CXCL10", "CXCL11", "CD40", "CCL22", "HAVCR2", "HLA-DRB5", "GZMB", "IGF1R", "CSF1R", "DAB2"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage_pDC_subcluster/pDC-diag_vs_rel",
         comp_grps = list(c("diagnostic", "relapse")),
         cluster_or_cell_types = c("pDC"),
         width=20, height=10)

# diagnostic vs. early relapse

# macrophage
VP_by_gene(data = hl.macrophage_pDC.subclustered,
           gene_list = c("CD163", "CD274", "IL3RA", "CD80", "CD86", "CXCL9", "CXCL10", "CXCL11", "CD40", "CCL22", "HAVCR2", "HLA-DRB5", "GZMB", "IGF1R", "CSF1R", "DAB2"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage_pDC_subcluster/macrophage-diag_vs_early_rel",
         comp_grps = list(c("diagnostic", "earlyRelapse")),
         cluster_or_cell_types = c("macrophage"),
         width=20, height=10, 
         earlyRelapse = TRUE)

# pDC
VP_by_gene(data = hl.macrophage_pDC.subclustered,
           gene_list = c("CD163", "CD274", "IL3RA", "CD80", "CD86", "CXCL9", "CXCL10", "CXCL11", "CD40", "CCL22", "HAVCR2", "HLA-DRB5", "GZMB", "IGF1R", "CSF1R", "DAB2"),
         file_name = "~/Documents/sc_HL_relapse/seurat/figures/V4_LATEST/vln_plots/macrophage_pDC_subcluster/pDC-diag_vs_early_rel",
         comp_grps = list(c("diagnostic", "earlyRelapse")),
         cluster_or_cell_types = c("pDC"),
         width=20, height=10, 
         earlyRelapse = TRUE)


```

# Generate violin plots for a given set of genes looking at a specific cell type / cluster and comparison

Parameters:
  - data: Seurat object with identities set to appropriate cell types or clusters
  - file_name: file prefix (including folder path)
  - comp_grps: list of DE groups to include
  - cluster_or_cell_types: list of cell types (or clusters) to plot data for
  - width: width of plot
  - height: height of plot

```{r}
VP_by_gene <- function(data,
                       gene_list, 
                       file_name, 
                       comp_grps, 
                       cluster_or_cell_types, 
                       width, 
                       height, 
                       earlyRelapse = FALSE){
  
  data.df <- NULL
  expr.df <- NULL

  plot_one_gene <- function(gene){
    
    # Step 1: create expr.df dataset (contains columns: expr, case, celltype/cluster, mean_expr)
      
    # Get the expression data
    if (earlyRelapse) {
      data.df <- data.frame(expr = data[["RNA"]]@data[gene,], 
                              case = data$earlyRelapse, 
                              celltype = data$celltype2)
    } else {
      data.df <- data.frame(expr = data[["RNA"]]@data[gene,], 
                              case = data$case.type, 
                              celltype = data$celltype2)
    }
    expr.df <- data.df %>% filter(celltype %in% cluster_or_cell_types, case %in% comp_grps[[1]])

    # Step 2: Calculate mean / median
    expr.df <- as.data.frame(expr.df %>%
        select(expr, case) %>%
        group_by(case) %>%
        mutate(mean_expr = mean(expr)))

    # Step 3: Plot (note that we are using the Wilcoxon-rank-sum test (deals with smaller numbers better))
    ggplot(expr.df, aes(x = case, y = expr)) + 
      geom_violin(aes(fill = mean_expr), trim = TRUE, scale = "width") +
      #scale_y_continuous(trans='log2') +
      # Add 10% space on the y-axis above the box plots
      scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
      stat_compare_means(comparisons = comp_grps, label = "p.format") +
      stat_summary(fun.data = custom_quantile, geom = "boxplot", width=0.1) +
      xlab("") + ylab("") + ggtitle(gene) +
      theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) +
      theme(axis.text = element_text(size = 12))
    
  }
  purrr::map(gene_list, plot_one_gene) %>% cowplot::plot_grid(plotlist = .)
  file_name <- paste(file_name, ".png", sep = "")
  ggsave(file_name, width = width, height = height)
}
```

## Create heatmap of average expression of clusters across component markers

Parameters:
  - data: Seurat object with $seurat_clusters

```{r}
avgComponentExpressionHeatmap <- function(data) {
  # get component markers
  component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", 
                                   sep = "\t", header = TRUE, fill = TRUE)
  # first, get list of highly-variable genes
  hvg <- data@assays$integrated@var.features
  # only look at markers that are in the HVG list
  markers.to.plot <- subset(hvg, hvg %in% component.mappings$Gene)
  # get average expression of each gene
  cluster.averages <- AverageExpression(data, features = markers.to.plot, return.seurat = TRUE)
  data.avg <- GetAssayData(object = cluster.averages, slot = "scale.data")
  # now match component gene list to those being plotted
  component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
  rownames(component.mappings) <- component.mappings$Gene
  # order by component then gene name so heatmap legend makes sense
  components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
  component.mappings$Gene <- NULL
  component.mappings$Alt_name <- NULL
  # reorder expression matrix to match component mappings
  data.avg.ordered <- data.avg[match(rownames(components.ordered), rownames(data.avg)), ]

  # heatmap of average expression per cluster across all component genes
  p <- pheatmap(hl.subclustered.components.cluster.avgs.ordered, 
         annotation_row = component.mappings,
         cluster_rows = FALSE, 
         annotation_names_row = FALSE, 
         angle_col = 45, 
         gaps_row = c(11,17,19,22,27,29,33,34,35,36,39,46,48), border_color = NA)
  # rotated version of heatmap
  p.rotated <- pheatmap(t(hl.subclustered.components.cluster.avgs.ordered), 
         annotation_col = component.mappings,
         cluster_cols = FALSE, 
         annotation_names_col = FALSE, 
         gaps_col = c(11,17,19,22,27,29,33,34,35,36,39,46,48), 
         angle_col = 45, 
         border_color = NA)
}

```

## Produce single-cell level heatmaps of expression for EACH cluster

Parameters:
  - data: Seurat object with $seurat_clusters
  - plot_folder: absolute path of folder to print heatmaps to

```{r}
printSCHeatmapsPerCluster <- function(data, plot_folder) {
  # get component markers
  component.mappings <- read.table("~/Documents/sc_HL_relapse/data/component_genes_with_functions.txt", 
                                   sep = "\t", header = TRUE, fill = TRUE)
  # first, get list of highly-variable genes
  hvg <- data@assays$integrated@var.features
  # only look at markers that are in the HVG list
  markers.to.plot <- subset(hvg, hvg %in% component.mappings$Gene)
  # get total number of clusters
  NUM_CLUSTERS = length(unique(data$seurat_clusters))
  # get expression matrix (features by barcodes)
  data.expr <- data@assays$SCT@data
  # retrieve expression for each cell across the component genes of interest
  data.expr.markers <- data.expr[ rownames(data.expr) %in% markers.to.plot , ]
  # match component gene list to those that can be plotted
  component.mappings <- filter(component.mappings, component.mappings$Gene %in% markers.to.plot)
  # order by component then gene name so that the heatmap legend makes sense
  rownames(component.mappings) <- component.mappings$Gene
  components.ordered <- component.mappings[order(component.mappings$Component, component.mappings$Gene), ]
  components.ordered$Gene <- NULL
  components.ordered$Alt_name <- NULL
  # reorder expression matrix to match component mappings
  data.expr.markers.ordered <- data.expr.markers[ row.names(components.ordered), ]
  # to create separate expr matrices for each cluster, need mappings for cell --> cluster
  data.clusters <- data$seurat_clusters
  # convert to char
  clusters.char <- levels(data.clusters)[data.clusters] # ordered vector of cluster ids (0 through 26)
  barcodes.cluster <- attributes(data.clusters)$names # ordered vector of barcodes (76,607 unique entries)
  clusters.df <- data.frame(barcodes.cluster, clusters.char)
 
  expr.markers <- vector('list', NUM_CLUSTERS)
  heatmaps <- vector('list', NUM_CLUSTERS)
  hl.seurat <- vector('list', NUM_CLUSTERS)
  
  for (c in 1:NUM_CLUSTERS) {
    hl.seurat[[c]] <- subset(data, idents = c-1) # filter by cluster
    hl.seurat[[c]] <- subset(hl.seurat[[c]], features = markers.to.plot) # filter by genes
    
    # retrieve expression for each cell across the component genes of interest
    expr.temp <- hl.seurat[[c]]@assays$SCT@data
    expr.markers[[c]] <- expr.temp[ rownames(expr.temp) %in% markers.to.plot , ]
    
    # reorder expression matrix to match component mappings
    expr.markers[[c]] <- expr.markers[[c]][ row.names(components.ordered), ]
  
    filename <- paste(plot_folder, "heatmap-cluster", 
                      c-1, ".pdf", sep = "")
    pdf(filename, width = 8, height = 8)
    heatmaps[[c]] <- pheatmap(expr.markers[[c]], 
           annotation_row = components.ordered, 
           cluster_rows = FALSE, 
           annotation_names_row = FALSE, 
           angle_col = 45, 
           gaps_row = c(11,17,19,22,27,29,33,34,35,36,39,46,48), 
           border_color = NA, show_colnames = FALSE, scale = "none")
    dev.off()
  }
 
}
```

## Perform DE analysis on all pairs

Parameters:
  - data: Seurat object
  - grep_string: grep string representing comparators to exclude
  - file_name: name of file within working directory

```{r}
# prepare data for DE analysis, run the DE analysis, and write the results to output
writeDEresults <- function(data, grep_string, file_name ) {
  Idents(data) <- data$comparator
  # get the comparators
  comparators <- sort(unique(data$comparator))
  # store number of cells for each comparator
  cell_counts.df <- as.data.frame(table(data$comparator))
  # remove cases not of interest
  comparators <- comparators[ !grepl(grep_string, comparators)]
  # perform DE analysis, where odd indices = group1 (e.g. diagnostic) and even indices = group2 (e.g. relapse)
  de.results <- get_de_genes(data = data, 
                             comparators = comparators, 
                             de.test = "MAST", 
                             cell_counts.df = cell_counts.df)
  de.results.df <- merge_de_results(de.results)
  write.table(de.results.df, paste(work_dir, file_name, sep=""), sep = "\t", quote = FALSE, row.names = FALSE)
}

# function to find DE genes between all pairs
get_de_genes <- function(data, comparators, de.test, cell_counts.df) {
  
  # get list of comparators that have fewer than 100 cells
  cell_counts.df <- cell_counts.df[ cell_counts.df$Var1 %in% comparators, ] # only look at comparators of interest
  low_cell_counts.comparators <- cell_counts.df[ cell_counts.df$Freq < 3, ]$Var1
  
  NUM_CLUSTERS = length(comparators) / 2
  # de_results <- vector('list', NUM_CLUSTERS)
  de_results <- list()

  for (i in 1:NUM_CLUSTERS) {
    index.grp1 <- (i*2)-1
    index.grp2 <- i*2
    
    # check that both grp1 and grp2 are NOT in the low cell counts list
    if ( !(comparators[index.grp1] %in% low_cell_counts.comparators) && 
         !(comparators[index.grp2] %in% low_cell_counts.comparators)) {
      
        # slot = "data" (normalized counts)
        tempResults <- FindMarkers(data, assay = "RNA", slot = "data",
                                   ident.1 = comparators[index.grp1], 
                                   ident.2 = comparators[index.grp2], 
                                   test.use = de.test, 
                                   latent.vars = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
                                   min.pct = 0.25, 
                                   logfc.threshold = 0)
        # make gene a column
        tempResults$gene <- rownames(tempResults)
        # add comparison description as a column
        tempResults$comparison <- paste(comparators[index.grp1], "_VS_", comparators[index.grp2], sep = "")
        rownames(tempResults) <- NULL
        # change order of columns
        cols <- c("comparison", "gene", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")
        tempResults <- tempResults[ cols ]
        # sort by descending fold change
        tempResults <- tempResults[order(-tempResults$avg_log2FC), ]
        # filter out non-interesting genes (i.e. *orf*, ^RP, and ^MT- genes)
        tempResults <- tempResults[ !(grepl("(^RP)|(MT-)|(*orf*)", tempResults$gene)), ]
        # remove results that have an adjusted p-value of 1
        tempResults <- dplyr::filter(tempResults, tempResults$p_val_adj < 1)
        
        # !!! NB: this step will fail if tempResults is empty (e.g. all adjusted p-values are 1)
        # store the processed results in the final result array
        if (nrow(tempResults) > 0) {
          de_results[[unique(tempResults$comparison)]] <- tempResults  
        }
      }
  }
  return(de_results)
}

# function to merge results into single dataframe
merge_de_results <- function(results.vector) {
  # initialize dataframe
  merged.df <- data.frame(comparison = character(),
                          gene = character(),
                          p_val = double(),
                          avg_logFC = double(),
                          pct.1 = double(),
                          pct.2 = double(),
                          p_val_adj = double())
  
  NUM_CLUSTERS = length(results.vector)
  for (i in 1:NUM_CLUSTERS) {
    merged.df <- rbind(merged.df, results.vector[[i]])
  }
  return(merged.df)
}
```

